---
title: "An Example of Simulating a Trial with Adaptive Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Example of Simulating a Trial with Adaptive Design}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  cache.path = 'cache/adaptiveDesign/',
  comment = '#>',
  dpi = 300,
  out.width = '100%'
)
```

```{r setup, echo = FALSE, message = FALSE}
library(dplyr)
library(R6)
library(rlang)
library(survival)
library(survminer)
library(ggplot2)
library(TrialSimulator)
```

In this vignette, we illustrate how to use `TrialSimulator` to simulate
a trial with seamless adaptive design, where dose selection, futility
analysis, interim analysis are triggered by user-defined events.

First we define endpoints in the trial. User can define arbitrary number
of endpoints of class `Endpoint` using the `new` function in `R6`. PFS
and OS are used as endpoints in this trial. In phase II, a binary
surrogate is used for dose selection. We consider two active arms (low
or high dose) and a placebo arm with `1:1:1` randomization. In the
placebo arm, the median of PFS and OS are 5 and 14 months, respectively.
The binary endpoint `surrogate` is of a rate 0.02. By default, the
permuted block randomization is executed.

```{r ljgai, class.source="fold-show", dpi = 1200}
#' define endpoints in arm <pbo>
pfs <- Endpoint$new(name = 'pfs', type='tte', 
                    generator=rexp, rate = log(2) / 5)
#' test the random number generator with a method of the class Endpoint
#' Note that user does not need to run this to set up a trial
#' This is for illustration purpose only
pfs_test <- pfs$test_generator(10000)
head(pfs_test)
km <- survfit(Surv(pfs, pfs_event) ~ 1, data = pfs_test)
ggsurvplot(km, data = pfs_test, legend = 'none', ggtheme = theme_minimal())
```

Similarly, we can define OS and surrogate for placebo. Here built-in
functions `rexp` and `rbinom` are used. User-defined functions are also
supported as long as it has the first argument `n`, number of samples to
be drawn, and returns a data frame of `n` rows. For any time-to-event
endpoint (e.g., `os`) in the data frame, `TrialSimulator` expects a
column ending with `_event` (e.g., `os_event`) being present as the
event indicator. In general, user can define their own generator that
returns more than one endpoints. For example, correlated endpoints would
be generated together, using packages i.e., `simstudy`, `gestate`,
`simdata`, etc. Note that `surrogate` is not a time-to-event endpoint,
so a readout time (`surrogate_readout`) is also needed when creating the
endpoint. In this example, the readout time is 2 weeks. Once all
endpoints are defined, an arm can be defined using class `Arm` and its
method `new`.

```{r dfaffageaf, class.source="fold-show"}
#' define a time-to-event endpoint OS, 
#' exponential distribution with median 13.71 months
os <- Endpoint$new(name = 'os', type='tte', 
                   generator = rexp, rate = log(2)/14)

head(os$test_generator(1000))
median(os$test_generator(1000)$os)

#' define binary endpoint VRR, with readout time = 2 weeks
two_weeks <- 2 / 52 * 12
surrogate <- Endpoint$new(
  name = 'surrogate', type = 'binary', readout = c(surrogate = two_weeks), 
  generator = rbinom, size = 1, prob = .02)

head(surrogate$test_generator(1000))
mean(surrogate$test_generator(1000)$surrogate)

#' define the placebo arm
placebo <- Arm$new(name = 'pbo', description = 'placebo')
#' add endpoints into the arm
placebo$add_endpoints(pfs, os, surrogate)
```

Likewise, we can define the three endpoints in the two active arms.

```{r}
#' define endpoints in the low dose arm
#' hazard ratios 
hr <- c(pfs_low = 0.7, os_low = 0.8, pfs_high = 0.65, os_high = 0.75)

pfs <- Endpoint$new(name = 'pfs', type='tte', 
                    generator = rexp, rate = log(2)/5 * hr['pfs_low'])

surrogate <- Endpoint$new(
  name = 'surrogate', type = 'binary', readout = c(surrogate = two_weeks), 
  generator = rbinom, size = 1, prob = .1) # lower probability

os <- Endpoint$new(name = 'os', type='tte', 
                   generator=rexp, rate = log(2)/14 * hr['os_low'])

low <- Arm$new(name = 'low dose', description = 'treated with low dose')
low$add_endpoints(pfs, os, surrogate)

#' define endpoints in the high dose arm
pfs <- Endpoint$new(name = 'pfs', type='tte', 
                    generator = rexp, rate = log(2)/5 * hr['pfs_high'])

surrogate <- Endpoint$new(
  name = 'surrogate', type = 'binary', readout = c(surrogate = two_weeks), 
  generator = rbinom, size = 1, prob = .12) # higher probability

os <- Endpoint$new(name = 'os', type='tte', 
                   generator=rexp, rate = log(2)/14 * hr['os_high'])

high <- Arm$new(name = 'high dose', description = 'treated with high dose')
high$add_endpoints(pfs, os, surrogate)
```

With three arms, we can define a trial of class `Trial`, where up to
1000 patients are enrolled. Recruitment curve are specified through
`enroller` with a built-in function `StaggeredRecruiter` of piecewise
constant rate: 20 patients per month until 15 months, and then 25
patients per month until the end of the trial. built-in random number
functions like `rexp` and user-defined function are also accepted.
Sample ratio can be specified through `sample_ratio`, which can be
updated during a trial using method `update_sample_ratio`. We set a
dropout rate 25% per 12 months. By default, no dropout if users do not
specify one.

```{r lagieg, class.source="fold-show"}
accrual_rate <- data.frame(end_time = c(15, Inf), piecewise_rate = c(20, 25))
trial <- Trial$new(
  name = 'Trial-3415', n_patients = 1000,
  seed = NULL, duration = 50,
  enroller = StaggeredRecruiter, accrual_rate = accrual_rate, 
  dropout = rexp, rate = -log(1 - 0.25)/12
)

trial$add_arms(sample_ratio = c(1, 1, 1), low, high, placebo)
```

The class `Trial` maintains a clock (`trial$get_current_time()`). Data
of patients who are enrolled before the time point are considered locked
and cannot be changed. Data of patients who is not yet recruited will be
regenerated immediately whenever the setting of trial are updated. Note
that randomization may be re-executed. Trial's setting is considered
changed if an arm is added (method `add_arms`), removed (method
`remove_arms`) or the sample ratio of one or multiple arm(s) is updated
(method `update_sample_ratio`). These methods are usually invoked in
user-defined action function when pre-specified events are triggered,
which will be discussed later.

The `Trial` only manages its arms and endpoints. In `TrialSimulator`, a
`Listener` is used to launch and monitor a trial according to a set of
`Event` that register to it. An event consists two main parts, one is
the condition to trigger it, one is the action function defining the
action to be taken once the condition is met. Usually, a condition is
met if a pre-specified number of events (or samples) are observed, or a
specific calendar time is reached. In this example, four events are
defined for dose selection, futility, interim and final analysis.

Dose selection and a non-binding futility analysis are planned to be
done at month 12. A built-in function `TriggerByCalendarTime` is used as
the condition function. Once data is locked, and trial's clock is
updated. A user-defined action function, `dose_selection_action` will be
executed by the listener. It is recommended that the user should load
the locked data at current time in the action function. Analysis is then
performed on it to support further actions. In this example, an active
arm can be dropped, and sample ratio of the remaining arms are updated
accordingly. All not yet enrolled patients' data is then regenerated
afterward. `Trial` has a public member function `save` for user to save
results for analysis or summary later. Note that `save` will be invoked
automatically when an event is triggered to save event time.

```{r aotel, class.source="fold-show"}
#' action function is defined by user. 
#' it always take at least two arguments, 
#' (1) trial: an object of class Trial
#' (2) event_name: character of event's name
#' It is recommended that user should always requests for the locked data
#' at current time when the event is triggered, from which the user can 
#' carry out various analysis fitting their purpose
dose_selection_action <- function(trial, event_name){

  locked_data <- trial$get_locked_data(event_name)
  ## do something with locked_data, e.g. estimate, Dunnett test, etc.,
  ## to support further actions
  ## ...

  ## this is just for illustration
  ## pick an arm to drop
  removed_arm <- sample(c('high dose', 'low dose', 'none'), size = 1, prob = c(.01, .01, .98))

  ## action 1: remove an arm <affected_arm>
  if(removed_arm == 'none'){
    return(NULL)
  }
  
  trial$remove_arms(removed_arm)

  ## action 2: update sample ratio for the remaining arms
  trial$update_sample_ratio('pbo', 1)
  if(removed_arm == 'high dose'){
    ## if low dose arm is kept, increase its ratio
    trial$update_sample_ratio('low dose', 2)
  }else{
    ## if high dose arm is kept, reduce its ratio for safety
    trial$update_sample_ratio('high dose', 1)
  }

  # action 3: non-binding futility analysis
  z_pfs <- rnorm(1)
  
  ## save results, which can be accessed later using trial$get_output()
  trial$save(value = removed_arm, name = 'removed_arm')
  trial$save(value = z_pfs, name = 'z_futility')
  
  ## Not necessary because event time is saved automatically when the event is triggered
  # trial$save(value = trial$get_event_time(event_name), name = 'futility_time')

  ## return for displaying purpose (print on screen later)
  ## user can return NULL
  action <- data.frame(
    order = c(1, 2, 3),
    type = c('remove', 'update sample ratio', 'non-binding futility'),
    value = c(removed_arm,
              ifelse(removed_arm == 'high dose', 'low dose -> 2', 'high dose -> 1'),
              paste0('z (pfs) = ', round(z_pfs, 2)))
  )

  action

}

## dose selection is carried out at month 12
dose_selection_event <- Event$new(name = 'dose selection', type = 'dose selection',
                                  trigger_condition = TriggerByCalendarTime,
                                  action = dose_selection_action,
                                  calendar_time = 12 + two_weeks)
```

Likewise, we can define events for interim and final analyses of PFS and
OS. The interim analysis for PFS will be carried out when 150 events are
observed in remaining arms.

```{r}
#' event to trigger interim analysis for PFS
pfs_interim_action <- function(trial, event_name){

  locked_data <- trial$get_locked_data(event_name)

  ## this p-value is just for illustration
  ## it can be calculated from locked_data
  pfs_interim_pval <- runif(1, 1e-3, .05)
  
  trial$save(value = pfs_interim_pval, name = 'pfs_interim_pval')

  ## action 1: if pfs is significant, increase its sample ratio
  ##           and reduce sample ratio of pbo.
  ##           otherwise no action.
  action <- paste0('No action. P-value: <pfs interim = ',
                   signif(pfs_interim_pval, 2), '>. ')
  if(pfs_interim_pval < .05){
    active_arm <- setdiff(trial$get_arms_name(), 'pbo')[1]
    trial$update_sample_ratio(active_arm, 2)
    trial$update_sample_ratio('pbo', 1)

    ## for displaying purpose, not essential
    action <- data.frame(
      order = c(1, 2),
      type = 'update sample ratio',
      value = c(paste0(active_arm, ' -> 2'), 'pbo -> 1')
    )
  }

  ## return for displaying purpose
  action

}

pfs_interim_event <- Event$new(name = 'pfs interim', type = 'interim analysis',
                            trigger_condition = TriggerByEventNumbers,
                            action = pfs_interim_action,
                            endpoints = 'pfs',
                            target_n_events = 150)
```

The final analysis for PFS will be performed with an interim of OS when
400 events are observed for PFS, or 250 events for OS, whichever comes
first.

```{r}
#'  event to trigger final analysis for PFS and interim analysis for OS
#'  In this example, the triggering condition is defined by number of events 
#'  of PFS and OS together. See meet = 'any' below.
pfs_final_action <- function(trial, event_name){

  locked_data <- trial$get_locked_data(event_name)
  ## do something with locked_data, e.g. estimate, test, etc.,
  ## to support further actions
  ## ...

  ## this is just for illustration
  pfs_final_pval <- runif(1, 0, 0.1)
  os_interim_pval <- runif(1, 0, 0.3)
  
  trial$save(
    data.frame(pfs_final_pval=pfs_final_pval, os_interim_pval=os_interim_pval)
  )

  ## only compute p-values for pfs and os, no further action
  action <- paste0('No action. P-values: <pfs final = ',
                   signif(pfs_final_pval, 2),
                   '>, <os interim = ', signif(os_interim_pval, 2), '>.')

  ## return for displaying purpose
  action

}
## meet = 'any' means that final analysis for PFS is triggered
## when pfs event reaches 400 or os event reaches 250.
## if meet = 'all', then final analysis for PFS is carried out only if
## both conditions are met.
pfs_final_event <- Event$new(name = 'pfs final', type = 'pfs final/os interim',
                               trigger_condition = TriggerByEventNumbers,
                               action = pfs_final_action,
                               endpoints = c('pfs', 'os'),
                               target_n_events = c(400, 250), meet = 'any')
```

Finally, one more analysis will be done for OS only at month 64.

```{r}
#'  event to trigger final analysis OS
#'  triggering condition: month 50
os_final_action <- function(trial, event_name){

  locked_data <- trial$get_locked_data(event_name)
  ## do something with locked_data, e.g. estimate, test, etc.,
  ## to support further actions
  ## ...

  ## this is just for illustration
  os_final_pval <- runif(1, 1e-3, .1)
  
  trial$save(os_final_pval, 'os_final_pval')

  ## action 1: end the trial, no futher action
  action <- paste0('No action. P-value: <os final = ',
                   signif(os_final_pval, 2), '>. ')

  ## return for displaying purpose
  action

}

os_final_event <- Event$new(name = 'os final', type = 'os final analysis',
                            trigger_condition = TriggerByCalendarTime,
                            action = os_final_action,
                            calendar_time = 50)
```

Now we can register the four events with the listener, and launch the
trial.

```{r aloganl, class.source="fold-show"}
listener <- Listener$new()
#' register events with listener
listener$add_events(
  dose_selection_event,
  pfs_interim_event,
  pfs_final_event,
  os_final_event
)

controller <- Controller$new(trial, listener)
controller$run(silent = FALSE)
```

In the action functions, user can save results to the `Trial` object
using its public member function `save` or any place they chose so that
they can summarize the trial later. The package can also summarize
commonly used metrics by default. For common design, the package can
carry out standard analysis, e.g., group sequential design with
combination test or graphical test as well, which would be implemented
and tested later.

```{r alojfqoitl, class.source="fold-show"}
trial$get_output()
```

In summary, `TrialSimulator` abstracts the data generation and
management to allow user focus on implementing the statistical analysis.
It simulates a trial on patient level, and provide flexibility in
adaptive design.
