---
title: "An Example of Simulating a Trial with Adaptive Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Example of Simulating a Trial with Adaptive Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  cache.path = 'cache/adaptiveDesign/',
  comment = '#>'#,
  #out.width = '80%'
)
```

```{r setup, echo = FALSE}
library(dplyr)
library(R6)
library(rlang)
library(survival)
library(survminer)
library(ggplot2)
library(TrialSimulator)
```

In this vignette, we illustrate how to use `TrialSimulator` to develop codes 
to simulate a trial with seamless design, where dose selection, futility 
analysis, interim analysis are triggered by user-defined events. 

First we define endpoints in the trial. User can define arbitrary number of 
endpoints of class `Endpoint` using the `new` function in `R6`. 
PFS and OS are used as endpoints in this trial. In phase II, 
a binary endpoint, VRR, is used in dose selection. We consider two active arms 
(low or high dose) and a placebo arm with `1:1:1` randomization. By default, 
the permuted block randomization is executed. For PFS, we assume a piecewise 
constant exponential distribution in the placebo arm. Here we use a builtin 
function (`TrialSimulator::PiecewiseConstantExponentialRNG`) to specify the random number generator for PFS.

```{r ljgai, class.source="fold-show"}
#' define endpoints in arm <pbo>
#' 
#' define a time-to-event endpoint PFS, with piecewise constant hazard
risk <- data.frame(
  end_time = c(1, 10, 26.0, 52.0),
  piecewise_risk = c(1, 1.01, 0.381, 0.150) * exp(-3.01)
)
pfs <- Endpoint$new(name = 'pfs', type='tte', 
                    generator=PiecewiseConstantExponentialRNG,
                    risk = risk, endpoint_name = 'pfs')
#'                     
#' test the random number generator with a method of the class Endpoint
pfs_test <- pfs$test_generator(10000)
head(pfs_test)
```
```{r}
km <- survfit(Surv(pfs, pfs_event) ~ 1, data = pfs_test)

ggsurvplot(km, data = pfs_test, legend = 'none', ggtheme = theme_minimal())
```

Similarly, we can define OS and VRR for placebo. Here builtin functions `rexp` 
and `rbinom` are used. User-defined functions are also supported as long as 
it has the first argument `n`, number of samples to be drawn, and returns a 
data frame of `n` rows. For any time-to-event endpoint (e.g., `ed`) in the data 
frame, `TrialSimulator` expects a column ending with `_event` (e.g., 
`ed_event`) being present as the event indicator. In general, user can define 
their own generator that returns more than one endpoints. For example, correlated 
endpoints would be generated together, using packages i.e., `simstudy`, 
`gestate`, `simdata`, etc. Note that VRR is not a time-to-event endpoint, 
so a readout time is also needed when creating the endpoint. 
Once all endpoints are defined, an arm can be created 
using class `Arm` and its method `new`. 
```{r dfaffageaf, class.source="fold-show"}
#' define a time-to-event endpoint OS, 
#' exponential distribution with median 25
os <- Endpoint$new(name = 'os', type='tte', generator = rexp, rate = log(2)/25)
head(os$test_generator(1000))
median(os$test_generator(1000)$os)

#' define binary endpoint VRR, with readout time = 2
vrr <- Endpoint$new(
  name = 'vrr', type = 'binary', 
  readout = c(vrr=2), generator = rbinom,
  size = 1, prob = .4)
head(vrr$test_generator(1000))
mean(vrr$test_generator(1000)$vrr)

#' define the placebo arm
placebo <- Arm$new(
  name = 'pbo', description = 'Placebo arm')
#' add endpoints into the arm
placebo$add_endpoints(pfs, os, vrr)
```

Likewise, we can define the three endpoints in the two active arms. 

```{r}
#' define endpoints in the low dose arm
risk$hazard_ratio <- .9 # HR
pfs <- Endpoint$new(name = 'pfs', type='tte', 
                    generator = PiecewiseConstantExponentialRNG,
                    risk = risk, endpoint_name = 'pfs')
vrr <- Endpoint$new(
  name = 'vrr', type = 'binary', 
  readout = c(vrr=2), generator = rbinom,
  size = 1, prob = .45) # slightly higher probability

os <- Endpoint$new(name = 'os', type='tte', 
                   generator=rexp, rate = log(2)/25 * .85) # HR = .85

low <- Arm$new(name = 'low dose', description = 'Low dose arm')
low$add_endpoints(pfs, os, vrr)

#' define endpoints in the high dose arm
risk$hazard_ratio <- .7
pfs <- Endpoint$new(name = 'pfs', type='tte', 
                    generator = PiecewiseConstantExponentialRNG, 
                    risk = risk, endpoint_name = 'pfs')

vrr <- Endpoint$new(
  name = 'vrr', type = 'binary', 
  readout = c(vrr=2), generator = rbinom, 
  size = 1, prob = .55) # higher probability

os <- Endpoint$new(name = 'os', type='tte', 
                   generator=rexp, rate = log(2)/25 * .75)

high <- Arm$new(name = 'high dose', description = 'High dose arm')
high$add_endpoints(pfs, os, vrr)
```

With three arms, we can define a trial of class `Trial`, where up to 1000 
patients are enrolled. Recruitment are modeled by exponential distribution. 
We also provide a builtin function `StaggeredRecruiter` for patient enrollment 
with piecewise constant rate. User-defined function are also supported. 
Sample ratio can be specified through `sample_ratio`, which can be updated 
during a trial using method `update_sample_ratio`. 

```{r lagieg, class.source="fold-show"}
n_patients <- 1000
trial <- Trial$new(
  name = 'Trial-3415', n_patients = n_patients,
  seed = 31415926,
  enroller = rexp, rate = log(2) / 5)

trial$add_arms(sample_ratio = 1, placebo)
trial$add_arms(sample_ratio = c(1, 1), enforce = TRUE, low, high)
#' or equivalent, one can add three arms together
#' trial$add_arms(sample_ratio = c(1, 1, 1), 
#'                placebo, low, high)
```

The class `Trial` maintains a clock (`trial$get_current_time()`). Data of 
patients who are enrolled before the time point are considered locked and cannot 
be changed. Data of patients who is currently unenrolled will be regenerated 
immediately whenever the setting of trial are changed. Note that randomization 
may be re-executed. Trial's setting is considered changed if an arm is added
(method `add_arms`), removed (method `remove_arms`) or the sample ratio of one 
or multiple arm(s) is updated (method `update_sample_ratio`). These methods are 
usually invoked after pre-specified events are triggered, which will be 
discussed later. 

The `Trial` only manages its arms and endpoints. A `Listener` will be used to 
launch and monitor a trial according to a set of `Event` that register to it. 
An event consists two main parts, one is the condition to trigger it, one is 
the action function defining the action to be taken once the condition is met. 
Usually, a condition is met if a pre-specified number of events (or samples) 
are observed, or a specific calendar time is reached. In this example, four 
events are defined for dose selection, futility, interim and final analysis. 

Dose selection is planned to be done when VRR are observed for 50 patients. 
A builtin function `TriggerByEventNumbers` is used as the condition function 
to define the event. The function can figure out the time point when this 
condition is met. Data is then locked, and the clock will be set to update the 
"current time" of the trial. A user-defined function, `futility_action` will 
be executed by the listener. In `futility_action`, it is recommended that 
the user should load the locked data at current time. Analysis is then done 
to support further actions. In this example, an active arm can be dropped, 
and sample ratio of the remaining arms are updated accordingly. All unenrolled
patients will have their data being regenerated afterward.

```{r aotel, class.source="fold-show"}
#' action function is defined by user. 
#' it always take at least two arguments, 
#' (1) trial: an object of class Trial
#' (2) event_name: character of event's name
#' It is recommended that user should always requests for the locked data
#' at current time when the event is triggered, from which the user can 
#' carry out various analysis fitting their purpose
futility_action <- function(trial, event_name){

  locked_data <- trial$get_locked_data(event_name)
  ## do something with locked_data, e.g. estimate, test, etc.,
  ## to support further actions
  ## ...

  ## this is just for illustration
  ## pick an arm to drop
  affected_arm <- sample(c('high dose', 'low dose'), size = 1, prob = c(.45, .55))

  ## action 1: remove an arm <affected_arm>
  trial$remove_arms(affected_arm)

  ## action 2: update sample ratio for the remaining arms
  trial$update_sample_ratio('pbo', 1)
  if(affected_arm == 'high dose'){
    ## if low dose arm is kept, increase its ratio
    trial$update_sample_ratio('low dose', 2)
  }else{
    ## if high dose arm is kept, reduce its ratio for safety
    trial$update_sample_ratio('high dose', 1)
  }

  ## return for displaying purpose (print on screen later)
  ## user can return NULL
  action <- data.frame(
    order = c(1, 2),
    type = c('remove', 'update sample ratio'),
    value = c(affected_arm, ifelse(affected_arm == 'high dose', 'low dose -> 2', 'high dose -> 1'))
  )

  action

}

futility_event <- Event$new(name = 'futility', type = 'futility analysis',
                    trigger_condition = TriggerByEventNumbers,
                    action = futility_action,
                    endpoints = 'vrr',
                    target_n_events = 150)
```

Likewise, we can define events for interim and final analyses of PFS and OS.

```{r}
#' event to trigger interim analysis for PFS
pfs_interim_action <- function(trial, event_name){

  locked_data <- trial$get_locked_data(event_name)
  ## do something with locked_data, e.g. estimate, test, etc.,
  ## to support further actions
  ## ...

  ## this p-value is just for illustration
  ## it can be calculated from locked_data
  pfs_interim_pval <- runif(1, 1e-3, .05)

  ## action 1: if pfs is significant, increase its sample ratio
  ##           and reduce sample ratio of pbo.
  ##           otherwise no action.
  action <- paste0('No action. P-value: <pfs interim = ',
                   signif(pfs_interim_pval, 2), '>. ')
  if(pfs_interim_pval < .05){
    active_arm <- setdiff(trial$get_arms_name(), 'pbo')
    trial$update_sample_ratio(active_arm, 2)
    trial$update_sample_ratio('pbo', 1)

    ## for displaying purpose, not essential
    action <- data.frame(
      order = c(1, 2),
      type = 'update sample ratio',
      value = c(paste0(active_arm, ' -> 2'), 'pbo -> 1')
    )
  }

  ## return for displaying purpose
  action

}

pfs_interim_event <- Event$new(name = 'pfs interim', type = 'interim analysis',
                            trigger_condition = TriggerByEventNumbers,
                            action = pfs_interim_action,
                            endpoints = 'pfs',
                            target_n_events = 400)
```
```{r}
#'  event to trigger final analysis for PFS and interim analysis for OS
#'  In this example, the triggering condition is defined by number of events 
#'  of PFS and OS together. See meet = 'any' below.
pfs_final_action <- function(trial, event_name){

  locked_data <- trial$get_locked_data(event_name)
  ## do something with locked_data, e.g. estimate, test, etc.,
  ## to support further actions
  ## ...

  ## this is just for illustration
  pfs_final_pval <- runif(1, 0, 0.1)
  os_interim_pval <- runif(1, 0, 0.3)

  ## only compute p-values for pfs and os, no further action
  action <- paste0('No action. P-values: <pfs final = ',
                   signif(pfs_final_pval, 2),
                   '>, <os interim = ', signif(os_interim_pval, 2), '>.')

  ## return for displaying purpose
  action

}
## meet = 'any' means that final analysis for PFS is triggered
## when pfs event reaches 300 or os event reaches 100.
## if meet = 'all', then final analysis for PFS is carried out only if
## both conditions are met.
pfs_final_event <- Event$new(name = 'pfs final', type = 'pfs final/os interim',
                               trigger_condition = TriggerByEventNumbers,
                               action = pfs_final_action,
                               endpoints = c('pfs', 'os'),
                               target_n_events = c(600, 500), meet = 'any')
```

```{r}
#'  event to trigger final analysis OS
#'  triggering condition: all 1000 patients are enrolled and 
#'  number of OS event reaches 800, i.e., meet = 'all', see below
os_final_action <- function(trial, event_name){

  locked_data <- trial$get_locked_data(event_name)
  ## do something with locked_data, e.g. estimate, test, etc.,
  ## to support further actions
  ## ...

  ## this is just for illustration
  os_final_pval <- runif(1, 1e-3, .1)

  ## action 1: end the trial, no futher action
  action <- paste0('No action. P-value: <os final = ',
                   signif(os_final_pval, 2), '>. ')

  ## return for displaying purpose
  action

}

os_final_event <- Event$new(name = 'os final', type = 'os final analysis',
                            trigger_condition = TriggerByEventNumbers,
                            action = os_final_action,
                            endpoints = c('vrr', 'os'),
                            target_n_events = c(n_patients, 800),
                            meet = 'all')
```

Now we can register the four events with the listener, and launch the trial.

```{r aloganl, class.source="fold-show"}
listener <- Listener$new()
#' register events with listener
listener$add_events(
    futility_event,
    pfs_interim_event,
    pfs_final_event,
    os_final_event
  )

controller <- Controller$new(trial, listener)
controller$run()
```

In the action functions, user can save results to the `Trial` object or 
any place they want to summarize the trial later. The package can also summarize 
commonly used metrics by default. For common design, the package can carry out 
standard analysis, e.g., group sequential design with combination test or 
graphical test as well, which would be implemented later. 

In summary, `TrialSimulator` abstracts the data generation and management to 
allow user focus on the programming for statistical analysis. It simulates a 
trial on patient level, and provide flexibility in adaptive design. 
