---
title: "Conditional Power and Event-Number Reassessment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conditional Power and Event-Number Reassessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>',
  fig.width = 7,
  fig.height = 4.5,
  dpi = 120,
  out.width = '100%'
)
```

```{r setup, message = FALSE}
library(TrialSimulator)
```

This vignette explains the statistical model and implementation behind
`Trials$conditionalPower()` and
`Trials$eventNumberReestimationFromConditionalPower()` in detail, including
the canonical distribution, alternative-direction mapping, effect
assumptions, nonmonotone root search, and integer algorithm. Both methods
apply to a time-to-event comparison with one interim analysis and one final
analysis. They use the interim logrank statistic and event count stored in a
triggered milestone.

The mathematical and diagnostic examples first isolate particular features
of the two methods. The final worked program then places the calculation in a
milestone action function and runs it through the standard `TrialSimulator`
framework, following the same action-based pattern used for other adaptive
and enrichment designs.

Conditional power is a design calculation, not an automatic adaptation rule.
In particular, changing a final event target after an unblinded interim can
affect type I error. A promising-zone design must be pre-specified and its
complete decision rule must be justified and evaluated, usually by simulation.

## Statistical setting

Consider one treatment arm compared with placebo. Let

-   $d$ be the number of events observed on the two arms at interim;
-   $D>d$ be the total number of events at a future final analysis;
-   $z$ be the interim logrank z statistic;
-   $r$ be the treatment-to-placebo allocation ratio; and
-   $\omega=r/(1+r)^2$ be the Schoenfeld information per event.

The package uses the sign convention of `fitLogrank()`: $z>0$ corresponds to
an estimated treatment-to-placebo hazard ratio greater than one. Therefore,
when a hazard ratio below one is beneficial, use `alternative = "less"`.

It is convenient to map both alternatives to a lower-tail problem. Define

$$
z_* = \begin{cases}
z, & \text{alternative = "less"},\\
-z, & \text{alternative = "greater"},
\end{cases}
\qquad
\theta_* = \begin{cases}
\log(h), & \text{alternative = "less"},\\
-\log(h), & \text{alternative = "greater"},
\end{cases}
$$

where $h$ is an assumed treatment-to-placebo hazard ratio. On this common
scale, a negative $z_*$ or $\theta_*$ favors treatment. The final rejection
region is

$$
Z_D^* \leq c_\alpha, \qquad c_\alpha=\Phi^{-1}(\alpha).
$$

For `alternative = "greater"`, this is equivalent to comparing the original
statistic with the upper boundary $\Phi^{-1}(1-\alpha)$.

### What `alpha` means

The argument `alpha` is the **one-sided nominal level corresponding to the
planned final critical boundary**. It is not necessarily the total design
alpha, the cumulative alpha spent by the final look, or the incremental alpha
spent at that look. For example, if an upper-tail group-sequential design has
final critical value $c$, pass `alpha = 1 - pnorm(c)`.

## Conditional power

Write $t=d/D$ for the interim information fraction. Under the canonical
independent-increments model, conditional on the observed interim statistic,

$$
Z_D^* \mid Z_d^*=z_*
\sim N\left\{
\sqrt{t}\,z_* + \theta_*\sqrt{\omega D}(1-t),\;1-t
\right\}.
$$

It follows that the lower-tail conditional power is

$$
CP(D)=\Phi\left[
\frac{
c_\alpha-\sqrt{d/D}\,z_*
-\theta_*\sqrt{\omega D}(1-d/D)
}{\sqrt{1-d/D}}
\right].
$$

The implementation provides three choices for the future effect.

### A specified hazard ratio

For a positive numeric `effect = h`, the method uses
$\theta_* = \log(h)$ for `"less"` and $-\log(h)$ for `"greater"`. This
separates the assumed future effect from random variation in the interim
estimate. It also requires the allocation ratio recorded at the milestone so
that $\omega$ can be computed.

### The interim trend

For `effect = "trend"`, the interim estimate is extrapolated:

$$
\widehat\theta_* = \frac{z_*}{\sqrt{\omega d}}.
$$

Substitution cancels $\omega$ and gives

$$
CP_{\text{trend}}(D)=\Phi\left[
\frac{c_\alpha-z_*\sqrt{D/d}}{\sqrt{1-d/D}}
\right].
$$

This calculation is often useful as a sensitivity analysis, but it can be
optimistic after a favorable interim because it treats the noisy interim
estimate as the true future effect.

### The null effect

For `effect = "null"`, the remaining increment has zero drift:

$$
CP_{0}(D)=\Phi\left[
\frac{c_\alpha-\sqrt{d/D}\,z_*}{\sqrt{1-d/D}}
\right].
$$

This is the conditional type I error at the specified final event number, not
power under a beneficial alternative. The event-number reassessment method
does not accept `effect = "null"`; use `conditionalPower()` directly when this
quantity is wanted.

### What happens near the interim

`conditionalPower()` requires $D>d$. If the interim statistic has crossed the
final boundary, $CP(D)$ tends to one as $D$ approaches $d$ from above, but it
need not remain high as more events accrue. Future independent increments can
dilute the interim result. Reaching the *final* boundary at interim is also not
the same as reaching a pre-specified interim efficacy boundary and does not by
itself authorize stopping a real trial.

## Inverting conditional power to obtain an event number

The member method has the following signature:

```r
trial$eventNumberReestimationFromConditionalPower(
  milestone, formula, placebo, alternative, alpha, target_cp,
  effect, ..., D_cap = NULL
)
```

For target conditional power $\gamma$, the reassessment method searches for

$$
D_{\text{new}}=
\min\left\{D\in\mathbb{Z}:d<D\leq D_{\text{cap}},\;CP(D)\geq\gamma\right\}.
$$

The strict requirement $D>d$ represents a genuine future final analysis. It
also applies when the interim z statistic already reaches the final boundary.

The cap has the following semantics:

-   `D_cap = NULL`, the default, is converted to `Inf` for every comparison
    and requests an unbounded search;
-   if a finite solution exists in the requested range, the smallest solution
    is returned in `D`, its conditional power is returned in `achieved_cp`,
    and `target_reached = TRUE`;
-   if no finite solution exists in the requested range, `D` and
    `achieved_cp` are `NA`, and `target_reached = FALSE`. `D_cap` continues to
    report the requested cap.

Thus, a false `target_reached` does not mean the algorithm failed. It means
that the requested conditional power cannot be attained within the permitted
event budget, or at any finite event number when the search is unbounded.

## Why ordinary bisection is insufficient

Conditional power need not be monotone in $D$. This matters particularly when
a specified effect differs from the observed interim trend. A test based only
on $CP(D_{\text{cap}})$ can miss an earlier target crossing, and bisection over
the whole range can converge to a later crossing.

To expose the shape analytically, put

$$
y=\sqrt{D/d}>1,
$$

and define the lower-scale accumulated drift

$$
\eta=\begin{cases}
z_*, & \text{effect = "trend"},\\
\theta_*\sqrt{\omega d}, & \text{numeric effect}.
\end{cases}
$$

Because $\Phi$ is monotone, it is enough to study its probit argument

$$
g(y)=\frac{c_\alpha y-\eta y^2+\eta-z_*}{\sqrt{y^2-1}}.
$$

Its derivative has the sign of

$$
-\eta y^3+(z_*+\eta)y-c_\alpha.
$$

Consequently, every stationary point is obtained from a **cubic equation**,
for both `effect = "trend"` and a numeric effect. There is no need to solve a
quartic equation for the target crossings themselves.

### Integer search algorithm

For each treatment-placebo comparison, the implementation:

1.  maps `"greater"` to the common lower-tail scale;
2.  starts at the first admissible integer, $D=d+1$;
3.  solves the stationary cubic with `polyroot()` and keeps real roots $y>1$;
4.  maps each retained root to $D=dy^2$ and includes both its floor and
    ceiling as integer knots;
5.  searches the resulting monotone integer intervals from left to right;
6.  uses integer bisection within the first interval whose upper endpoint
    reaches the target.

For an infinite cap, all bounded stationary intervals are searched first. On
the final branch,

$$
\lim_{D\to\infty}CP(D)=
\begin{cases}
1, & \eta<0,\\
\alpha, & \eta=0,\\
0, & \eta>0.
\end{cases}
$$

This limit determines whether a later finite crossing can exist. When it can,
the implementation expands an upper bracket geometrically and finishes with
integer bisection. Since a cubic has at most three real roots, only a few
intervals are inspected; the event-number range is not scanned integer by
integer.

### An example with three target crossings

The following hypothetical lower-tail example has a weak beneficial assumed
effect that differs substantially from the favorable interim result.

```{r multiple-crossings}
cp_formula <- function(D, z, d, alpha, hazard_ratio, omega = 0.25){
  t <- d / D
  theta <- log(hazard_ratio)
  pnorm((qnorm(alpha) - sqrt(t) * z -
           theta * sqrt(omega * D) * (1 - t)) / sqrt(1 - t))
}

z_example <- -1.8
d_example <- 100
target_example <- 0.2
D_grid <- 101:10000
cp_grid <- cp_formula(D_grid, z = z_example, d = d_example,
                      alpha = 0.025, hazard_ratio = 0.98)

crossing_index <- which(diff(cp_grid >= target_example) != 0) + 1
D_grid[crossing_index]
```

The target is crossed near 107, crossed downward near 183, and crossed upward
again near 8675. The requested answer is 107, including when a finite cap such
as 5000 has conditional power below the target.

```{r multiple-crossings-table}
selected_D <- c(106, 107, 182, 183, 5000, 8674, 8675)
data.frame(
  D = selected_D,
  conditional_power = round(
    cp_formula(selected_D, z = z_example, d = d_example,
               alpha = 0.025, hazard_ratio = 0.98),
    4
  ),
  reaches_target = cp_formula(selected_D, z = z_example, d = d_example,
                              alpha = 0.025,
                              hazard_ratio = 0.98) >= target_example
)
```

```{r multiple-crossings-plot}
plot(D_grid, cp_grid, type = 'l', log = 'x',
     xlab = 'Final event number D (log scale)',
     ylab = 'Conditional power', ylim = c(0, 0.3))
abline(h = target_example, lty = 2, col = 'firebrick')
abline(v = c(107, 183, 8675), lty = 3, col = 'grey40')
```

## Worked `TrialSimulator` example

We now simulate one two-arm trial with an event-driven endpoint. A calendar
milestone provides the locked interim data used by both methods.

```{r define-trial}
pbo <- arm(name = 'pbo')
pbo$add_endpoints(
  endpoint(name = 'pfs', type = 'tte', generator = rexp,
           rate = log(2) / 10)
)

trt <- arm(name = 'trt')
trt$add_endpoints(
  endpoint(name = 'pfs', type = 'tte', generator = rexp,
           rate = log(2) / 14)
)

accrual <- data.frame(end_time = Inf, piecewise_rate = 30)
tr <- trial(name = 'cp-example', n_patients = 400, duration = 40,
            seed = 31416, enroller = StaggeredRecruiter,
            accrual_rate = accrual, silent = TRUE)
add_arms(tr, sample_ratio = c(1, 1), pbo, trt)

lstn <- listener(silent = TRUE)
lstn$add_milestones(
  milestone(name = 'interim', when = calendarTime(time = 15)),
  milestone(name = 'final', when = calendarTime(time = 40))
)
controller(tr, lstn)$run(n = 1, silent = TRUE, plot_event = FALSE)
```

The next subsections deliberately analyze this completed replicate directly,
outside a milestone action function. This is diagnostic code: it makes the
inputs and outputs easy to compare while demonstrating effect assumptions,
alternative directions, caps, and return values one feature at a time. It is
not the recommended structure for a simulation study. After these focused
examples, `promising_zone_action()` and a complete event-driven simulation
program show how the same methods are used inside the `TrialSimulator`
framework.

The methods call `fitLogrank()` internally. We show the fit explicitly here to
make the interim inputs visible. Likewise, the direct method calls that follow
are intentional teaching devices rather than actions that can adapt this
already completed trial.

```{r interim-fit}
interim_fit <- fitLogrank(
  Surv(pfs, pfs_event) ~ arm,
  placebo = 'pbo', data = tr$get_locked_data('interim'),
  alternative = 'less', tidy = FALSE
)
interim_fit[, c('arm', 'placebo', 'z', 'info')]
```

### Conditional power under different assumptions

Suppose the planned final analysis has 300 events and its one-sided final
boundary corresponds to nominal `alpha = 0.022`.

```{r compare-effects}
planned_D <- 300
final_alpha <- 0.022

cp_at <- function(effect){
  tr$conditionalPower(
    'interim', Surv(pfs, pfs_event) ~ arm,
    placebo = 'pbo', alternative = 'less',
    alpha = final_alpha, D = planned_D, effect = effect
  )$cp
}

data.frame(
  assumption = c('interim trend', 'null future drift', 'hazard ratio 0.75'),
  conditional_power = c(cp_at('trend'), cp_at('null'), cp_at(0.75))
)
```

The difference between the trend and specified-effect calculations is not a
software discrepancy: they answer different conditional questions. A robust
design should examine assumptions chosen before looking at the interim result.

The direction transformation can also be checked directly by swapping the
reference arm. The treatment-vs-placebo hazard ratio 0.75 becomes the
placebo-vs-treatment hazard ratio $1/0.75$, and `"less"` becomes `"greater"`.

```{r compare-directions}
cp_less <- tr$conditionalPower(
  'interim', Surv(pfs, pfs_event) ~ arm,
  placebo = 'pbo', alternative = 'less',
  alpha = final_alpha, D = planned_D, effect = 0.75
)

cp_greater <- tr$conditionalPower(
  'interim', Surv(pfs, pfs_event) ~ arm,
  placebo = 'trt', alternative = 'greater',
  alpha = final_alpha, D = planned_D, effect = 1 / 0.75
)

data.frame(
  formulation = c('trt vs pbo, less', 'pbo vs trt, greater'),
  z = c(cp_less$z, cp_greater$z),
  conditional_power = c(cp_less$cp, cp_greater$cp)
)
```

### Event-number reassessment and the cap

Under the assumed hazard ratio 0.75, request 90% conditional power. Comparing
two caps illustrates both possible values of `target_reached`; the smaller
cap remains recorded in `D_cap`, while `D` and `achieved_cp` are `NA` because
no solution exists through that cap.

```{r reassess-events}
reassess_500 <- tr$eventNumberReestimationFromConditionalPower(
  'interim', Surv(pfs, pfs_event) ~ arm,
  placebo = 'pbo', alternative = 'less',
  alpha = final_alpha, target_cp = 0.9,
  effect = 0.75, D_cap = 500
)

reassess_600 <- tr$eventNumberReestimationFromConditionalPower(
  'interim', Surv(pfs, pfs_event) ~ arm,
  placebo = 'pbo', alternative = 'less',
  alpha = final_alpha, target_cp = 0.9,
  effect = 0.75, D_cap = 600
)

rbind(cap_500 = reassess_500, cap_600 = reassess_600)[,
  c('d', 'D', 'D_cap', 'target_cp', 'achieved_cp', 'target_reached')]
```

Omitting `D_cap`, or passing `NULL`, requests the unbounded answer and is
equivalent here to the solution under cap 600.

```{r reassess-unbounded}
tr$eventNumberReestimationFromConditionalPower(
  'interim', Surv(pfs, pfs_event) ~ arm,
  placebo = 'pbo', alternative = 'less',
  alpha = final_alpha, target_cp = 0.9, effect = 0.75
)[, c('d', 'D', 'D_cap', 'achieved_cp', 'target_reached')]
```

## A promising-zone policy

A common design pattern classifies conditional power at the originally
planned final event number into zones. The following thresholds are purely
illustrative:

-   **unfavorable:** $CP(D_{\text{planned}})<0.30$;
-   **promising:** $0.30\leq CP(D_{\text{planned}})<0.90$;
-   **favorable:** $CP(D_{\text{planned}})\geq0.90$.

For the specified hazard ratio 0.75, the worked example lies in the promising
zone.

```{r promising-zone}
lower_cp <- 0.30
target_cp <- 0.90
cp_planned <- cp_at(0.75)

zone <- if(cp_planned < lower_cp){
  'unfavorable'
}else if(cp_planned < target_cp){
  'promising'
}else{
  'favorable'
}

data.frame(planned_D, cp_planned, zone,
           reassessed_D = reassess_600$D,
           target_reached = reassess_600$target_reached)
```

One possible pre-specified policy is to retain the original design in the
unfavorable and favorable zones, and increase the event target in the
promising zone subject to a cap. A different design could stop for futility in
the unfavorable zone. These choices are properties of the trial design, not
of the calculator.

The following function implements the increase-only rule at an interim
milestone. The final milestone must already be registered and must not yet
have triggered. The complete program near the end of this vignette registers
this action and executes the simulation.

```{r promising-zone-action}
promising_zone_action <- function(trial){
  planned_D <- 300
  lower_cp <- 0.30
  target_cp <- 0.90
  D_cap <- 600

  cp <- trial$conditionalPower(
    'interim', Surv(pfs, pfs_event) ~ arm,
    placebo = 'pbo', alternative = 'less',
    alpha = 0.022, D = planned_D, effect = 0.75
  )$cp

  if(cp < lower_cp){
    zone <- 'unfavorable'
    decision <- 'unfavorable: retain planned event number'
    selected_D <- planned_D
  }else if(cp < target_cp){
    zone <- 'promising'
    reassessment <- trial$eventNumberReestimationFromConditionalPower(
      'interim', Surv(pfs, pfs_event) ~ arm,
      placebo = 'pbo', alternative = 'less',
      alpha = 0.022, target_cp = target_cp,
      effect = 0.75, D_cap = D_cap
    )

    if(reassessment$target_reached && reassessment$D > planned_D){
      selected_D <- reassessment$D
      decision <- 'promising: reassess to target conditional power'
    }else{
      selected_D <- reassessment$D_cap
      decision <- if(reassessment$target_reached){
        'promising: use cap because the earliest solution is not an increase'
      }else{
        'promising: use cap because no solution was found through the cap'
      }
    }

  }else{
    zone <- 'favorable'
    decision <- 'favorable: retain the planned design'
    selected_D <- planned_D
  }

  if(selected_D > planned_D){
    trial$update_milestone(
      'final',
      when = eventNumber(endpoint = 'pfs', n = selected_D,
                         arms = c('pbo', 'trt'))
    )
  }

  selected_cp <- trial$conditionalPower(
    'interim', Surv(pfs, pfs_event) ~ arm,
    placebo = 'pbo', alternative = 'less',
    alpha = 0.022, D = selected_D, effect = 0.75
  )$cp

  trial$save(zone, 'promising_zone')
  trial$save(cp, 'conditional_power_at_planned_D')
  trial$save(selected_D, 'selected_final_event_number')
  trial$save(selected_cp, 'conditional_power_at_selected_D')
  trial$save(selected_cp >= target_cp, 'target_reached_at_selected_D')
  trial$save(decision, 'promising_zone_decision')
}
```

Many promising-zone designs permit an increase but not a decrease from the
originally planned event number. The reassessment method itself searches from
$d+1$, not from `planned_D`. Therefore, an increase-only policy must verify
that the returned `D` is at least `planned_D`. In a nonmonotone case, simply
replacing a smaller result by `planned_D` is not guaranteed to preserve the
target conditional power; the admissible search domain and action should be
pre-specified explicitly. The action above uses `D_cap` as its pre-specified
increase-only fallback and separately records whether conditional power at
that selected event number reaches the target.

## Multiple treatment arms

Both methods support several treatment-placebo comparisons. `D`, `alpha`,
`target_cp`, and a non-scalar finite `D_cap` are named by treatment arm. Names,
not vector positions, determine matching. For example:

```{r multiple-arms-syntax, eval = FALSE}
trial$conditionalPower(
  'interim', Surv(pfs, pfs_event) ~ arm,
  placebo = 'pbo', alternative = 'less',
  D = c(low = 350, high = 400),
  alpha = c(high = 0.01, low = 0.015),
  effect = 0.75
)

trial$eventNumberReestimationFromConditionalPower(
  'interim', Surv(pfs, pfs_event) ~ arm,
  placebo = 'pbo', alternative = 'less',
  alpha = c(low = 0.015, high = 0.01),
  target_cp = c(high = 0.90, low = 0.85),
  effect = 'trend',
  D_cap = c(low = 600, high = 700)
)
```

Event counts are pair-specific: each `d` and `D` counts events on placebo plus
one treatment arm. With a shared placebo, event counts across comparisons need
not sum to a trial-wide total.

## Assumptions and design checks

Before using these calculations in a design, check the following.

1.  The final statistic and boundary must match the canonical calculation.
    The allocation ratio should remain constant, and the compared treatment
    and placebo arms should be concurrently enrolled.
2.  `alpha` must represent the nominal final boundary, not an arbitrary
    overall alpha or final-stage alpha increment.
3.  The effect assumption should be pre-specified. Report sensitivity to the
    interim trend and clinically relevant fixed hazard ratios when useful.
4.  `D_cap` should reflect a feasible event budget. Inspect `target_reached`;
    when it is `FALSE`, both `D` and `achieved_cp` are `NA`.
5.  The operational rule must distinguish a final boundary from a formal
    interim stopping boundary.
6.  Simulate the **entire adaptive design** under the null and relevant
    alternatives. Unblinded sample-size or event-number changes can inflate
    type I error unless the testing procedure or the restricted adaptation
    rule has been justified.

`TrialSimulator` computes the stated conditional quantities but cannot verify
that a data-dependent adaptation is valid for a particular confirmatory trial.
If adaptation changes the final statistic, information weights, allocation, or
critical boundary, use an appropriate adaptive or combination-test analysis
rather than treating the original final test as unchanged.

## A complete promising-zone simulation

The earlier object `tr` was deliberately small and was used only to inspect a
single locked dataset. We now define a separate trial with enough patients and
follow-up to make every event target through `D_cap = 600` feasible. The
interim and final analyses are event driven. The interim action is
`promising_zone_action()` defined above; the final action performs the planned
logrank test and saves the result.

```{r promising-zone-final-action}
promising_zone_final_action <- function(trial){
  fit <- fitLogrank(
    Surv(pfs, pfs_event) ~ arm,
    placebo = 'pbo', data = trial$get_locked_data('final'),
    alternative = 'less', tidy = FALSE
  )

  trial$save(fit$info, 'observed_final_event_number')
  trial$save(fit$z, 'final_z')
  trial$save(fit$p, 'final_p')
  trial$save(fit$p <= 0.022, 'reject_final')
}
```

The following is a complete simulation program: arms, trial, milestone
actions, listener, and controller are all connected before execution. The
initial final milestone requests 300 events. When the interim action selects
a larger target, `update_milestone()` replaces that trigger for the current
replicate only; the listener restores the original 300-event trigger before
the next replicate.

```{r promising-zone-program}
sim_pbo <- arm(name = 'pbo')
sim_pbo$add_endpoints(
  endpoint(name = 'pfs', type = 'tte', generator = rexp,
           rate = log(2) / 10)
)

sim_trt <- arm(name = 'trt')
sim_trt$add_endpoints(
  endpoint(name = 'pfs', type = 'tte', generator = rexp,
           rate = 0.75 * log(2) / 10)
)

sim_accrual <- data.frame(end_time = Inf, piecewise_rate = 30)
sim_trial <- trial(
  name = 'promising-zone', n_patients = 700, duration = 120,
  seed = 20260831, enroller = StaggeredRecruiter,
  accrual_rate = sim_accrual, silent = TRUE
)
sim_trial$add_arms(sample_ratio = c(1, 1), sim_pbo, sim_trt)

sim_listener <- listener(silent = TRUE)
sim_listener$add_milestones(
  milestone(
    name = 'interim', action = promising_zone_action,
    when = eventNumber(endpoint = 'pfs', n = 150,
                       arms = c('pbo', 'trt'))
  ),
  milestone(
    name = 'final', action = promising_zone_final_action,
    when = eventNumber(endpoint = 'pfs', n = 300,
                       arms = c('pbo', 'trt'))
  )
)

sim_controller <- controller(sim_trial, sim_listener)
sim_controller$run(n = 1, silent = TRUE, plot_event = FALSE)
```

The saved columns show the complete path from the interim zone and selected
event target to the final test for this replicate.

```{r promising-zone-single-output}
single_output <- sim_controller$get_output(tidy = TRUE)
single_output[, c(
  'promising_zone', 'conditional_power_at_planned_D',
  'selected_final_event_number', 'conditional_power_at_selected_D',
  'target_reached_at_selected_D', 'observed_final_event_number',
  'final_p', 'reject_final'
)]
```

For operating characteristics, increase `n` after resetting the controller.
The 1000-replicate run is not executed while building the vignette. As in
other simulation vignettes in this package, its reproducible output is stored
internally and loaded in the next chunk.

```{r promising-zone-run, eval = FALSE}
sim_controller$reset()
sim_controller$run(n = 1000, silent = TRUE, plot_event = FALSE)
conditional_power_output <- sim_controller$get_output()
```

```{r promising-zone-load-output, echo = FALSE}
conditional_power_output <-
  TrialSimulator:::getConditionalPowerOutput()
```

Here are the first five replicates, restricted to the main decision and final
analysis columns.

```{r promising-zone-output-head}
output_columns <- c(
  'promising_zone', 'conditional_power_at_planned_D',
  'selected_final_event_number', 'conditional_power_at_selected_D',
  'target_reached_at_selected_D', 'observed_final_event_number',
  'final_p', 'reject_final'
)
conditional_power_output[seq_len(5), output_columns]
```

The zone frequencies and mean selected event number summarize how often the
rule adapts the final milestone when both the data-generating and
conditional-power hazard ratios are 0.75.

```{r promising-zone-zone-summary}
zone_levels <- c('unfavorable', 'promising', 'favorable')
zone <- factor(conditional_power_output$promising_zone,
               levels = zone_levels)
zone_count <- table(zone)
mean_D_by_zone <- tapply(
  conditional_power_output$selected_final_event_number,
  zone, mean
)

data.frame(
  zone = zone_levels,
  replicates = as.integer(zone_count),
  percent = round(100 * as.numeric(prop.table(zone_count)), 1),
  mean_selected_D = round(as.numeric(mean_D_by_zone), 1)
)
```

Finally, the overall summary connects adaptation frequency, attainment of the
conditional-power target at the selected event number, and the final testing
result. These values describe this illustrative data-generating scenario;
they do not establish type I error control for the adaptation rule.

```{r promising-zone-overall-summary}
data.frame(
  metric = c(
    'final event target increased',
    'target CP reached at selected D',
    'final null hypothesis rejected'
  ),
  percent = round(100 * c(
    mean(conditional_power_output$selected_final_event_number > 300),
    mean(conditional_power_output$target_reached_at_selected_D),
    mean(conditional_power_output$reject_final)
  ), 1)
)
```

## References

-   Chen, Y. H. J., DeMets, D. L., and Lan, K. K. G. (2004). Increasing the
    sample size when the unblinded interim result is promising. *Statistics in
    Medicine*, 23, 1023--1038.
    <https://doi.org/10.1002/sim.1688>
-   Cui, L., Hung, H. M. J., and Wang, S.-J. (1999). Modification of sample
    size in group sequential clinical trials. *Biometrics*, 55, 853--857.
    <https://doi.org/10.1111/j.0006-341X.1999.00853.x>
-   Mehta, C. R., and Pocock, S. J. (2011). Adaptive increase in sample size
    when interim results are promising: a practical guide with examples.
    *Statistics in Medicine*, 30, 3267--3284.
    <https://doi.org/10.1002/sim.4102>
