---
title: "Simulate Trials with Treatment Switching"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate Trials with Treatment Switching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache.path = 'cache/crossoverTreatmentSwitching/',
  comment = '#>',
  dpi = 300,
  out.width = '100%'
)
```

```{r setup, echo = FALSE, message = FALSE}
library(TrialSimulator)
library(mvtnorm)
library(dplyr)
library(kableExtra)
```

The `TrialSimulator` package can handle three types of crossover design.
This vignette demonstrates how to simulate a trial where all or a subset of patients are offered the option of treatment switching over time. For example, patients can switch to a higher dose if disease progresses. To provide such a flexibility in simulation, `TrialSimulator` support a `regime` object in which users can define the subset of patients who switch, the switching time, and the updated endpoints under new treatments. 





