---
title: "Modular Clinical Trial Simulation: Overview of Core Functions and Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modular Clinical Trial Simulation: Overview of Core Functions and Workflow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TrialSimulator)
```

`TrialSimulator` maintains a private field, `now`, representing the
current time in a running trial. This field is initialized to zero.
Additionally, the package includes a data frame, `trial_data`, which
stores simulated data of all planned patients given the setting by
`now`. Data collected (recruited) or observed (readouts, or
time-to-event) before `now` is considered final and fixed. Because
`TrialSimulator` aims to conduct simulation for adaptive clinical trial
design, data of patients recruited after `now` can be refreshed if
trial's setting is changed.

When trial settings change--such as adding or removing arms (`add_arms`
or `remove_arms`) or updating the sampling ratio for one or more arms
(`update_sample_ratio`)--the package invokes the `roll_back` function.
This function performs two key actions.

1.  Saving enrollment times: The `enroll_time` for patients who have not
    yet been enrolled by `now` is preserved, as recruitment curve is
    pre-defined and fixed.
2.  Deleting future patient data: Data corresponding to patients not yet
    enrolled is removed from `trial_data`.

After rolling back, the package updates the randomization process. It
uses `permuted_block_randomization` to refresh the
`randomization_queue`, ensuring proper randomization for patients who
have not yet been enrolled. Following this, the `enroll_patients` method
generates new data for these patients, reflecting the updated trial
settings. This ensures that `trial_data` remains consistent with the
trial's current configuration. This is critical to allow the package to
handle complex adaptive design.

The trial's progress is monitored by a listener, which tracks
pre-defined events. When the condition for a event is met, the `now`
field is updated to the time of the event, triggering a data lock. At
this point, the data is "frozen", creating a snapshot of `trial_data` as
of `now` in `locked_data`. This locked dataset is saved and can be
accessed by event's name. A locked dataset serves as the basis for
executing pre-defined action functions of an event, such as dose
selection, futility analysis, interim analysis or final analysis.
