---
title: "Dose-Ranging Study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dose-Ranging Study}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache.path = 'cache/doseRanging/',
  comment = '#>',
  dpi = 300,
  out.width = '100%'
)
```

```{r setup, echo = FALSE, message = FALSE}
library(TrialSimulator)
library(DoseFinding)
library(dplyr)
library(kableExtra)
set.seed(12345)
```

In this vignette, we implement simulation for a dose-ranging trial under 
adaptive design. We start from a placebo arm and a treatment arm with 
highest possible dose. At an interim, if we observe promising results, we add 
arms with doses in between to inform decision at final analysis.

For the sake of simplicity, we assume a binary endpoint of interest. The
response rates at doses 0 (placebo), 0.5, 1.5, 2.5, 4 (highest dose) are
given by a `sigEmax` model below, where the response rate is assume to
be 10% in placebo, and 25% in arm of the highest dose (dose = 4). For further 
background information, refer to [vignette](https://cran.r-project.org/package=DoseFinding/vignettes/binary_data.html) 
of the `DoseFinding` package. 

```{r flaf}
mods <- DoseFinding::Mods(sigEmax = rbind(c(1, 3)), 
                          placEff = log(.1/(1 - .1)), 
                          maxEff = log(.25/(1 - .25)) - log(.1/(1 - .1)),
                          doses = c(0, 0.5, 1.5, 2.5, 4))

DoseFinding::plotMods(mods, trafo = function(x) 1/(1+exp(-x)))

## response rates on curve
x <- DoseFinding::getResp(mods, doses = c(0, 0.5, 1.5, 2.5, 4))
1 / (1 + exp(-unclass(x)))
```

## Simulation Settings

In our simulation, we assume 

- the response rates of binary endpoint as

| Dose              | 0    | 0.5  | 1.5  | 2.5  | 4    |
|-------------------|------|------|------|------|------|
| Response Rate (%) | 10.0 | 11.2 | 20.8 | 24.1 | 25.0 |

- readout time is 1 week. 
- no dropout. 
- starting with a placebo arm and a treatment arm of the highest dose (= 4), 
with a 1:1 randomization ratio.
  - enroll 5 patients per week for 7 weeks.
- an interim is planed when we have collected 30 readouts in total from the 
two arms. 
  - if z statistic of odds ratio is greater than 1
    - add three arms of dose 0.5, 1.5, and 2.5 to trial
    - enroll 20 patients per week with a randomization ratio 1:2:2:2:1, until 
    in total 150 patients are enrolled in the trial
    - final analysis is performed when readouts are available for all enrolled patients. 
  - otherwise, stop the trial.
  
## Define Placebo and High Dose Arms

```{r}
ep <- endpoint(name = 'ep', type = 'non-tte', readout = c(ep = 1),
               generator = rbinom, size = 1, prob = 0.1)
pbo <- arm(name = 'dose = 0.0')
pbo$add_endpoints(ep)

ep <- endpoint(name = 'ep', type = 'non-tte', readout = c(ep = 1),
               generator = rbinom, size = 1, prob = 0.25)
trt4 <- arm(name = 'dose = 4.0')
trt4$add_endpoints(ep)
```

## Define a Trial

```{r}
accrual_rate <- data.frame(end_time = c(7, Inf),
                           piecewise_rate = c(5, 20))

trial <- trial(name = '123', n_patients = 150, duration = 14,
               enroller = StaggeredRecruiter, accrual_rate = accrual_rate, 
               silent = TRUE)

trial$add_arms(sample_ratio = c(1, 1), pbo, trt4)
trial
```

<!-- Hide this helper function and display it in Appendix for better typesetting. -->
```{r go_nogo, echo=FALSE}
go_nogo <- function(data){

  ## candidate models for MCPMod
  doses <- c(0, 0.5, 1.5, 2.5, 4)
  candidates <- Mods(emax = c(.25, 1),
                     sigEmax = rbind(c(1, 3), c(2.5, 4)),
                     betaMod = c(1.1, 1.1),
                     placEff = log(.1/(1 - .1)),
                     maxEff = log(.25/(1 - .25)) - log(.1/(1 - .1)),
                     doses = doses)

  fit <- glm(ep ~ factor(arm) + 0, data = data, family = binomial)
  mu_hat <- coef(fit)
  S_hat <- vcov(fit)

  ## multiple contrast test
  test <- DoseFinding::MCTtest(dose = doses,
                               mu_hat, S = S_hat,
                               models = candidates, type = "general")

  ## model averaging
  model <- DoseFinding::maFitMod(dose = doses,
                                 mu_hat, S = S_hat,
                                 models = c("emax", "sigEmax", "betaMod"))
  
  ## predict response rate per dose
  prd <- predict(model, summaryFct = median, doseSeq = doses)
  
  ## convert to scale of probability
  prd_rate <- 1 / (1 + exp(-prd))

  ## go/no-go rule: MCP test p-value < 0.05 and estimated effect > 10%
  ifelse(min(attr(test$tStat, 'pVal')) < .05 &
           max(prd_rate - prd_rate[1]) > .1, 'go', 'no-go')

}
```

<!-- Hide this action function and display it later for better typesetting. -->

```{r action_at_interim, echo=FALSE}
action_at_interim <- function(trial){

  ## get data snapshot
  locked_data <- trial$get_locked_data('interim')

  ## compare two arms
  ## Risk difference = response rate in high dose - response rate in placebo
  fit <- fitLogistic(ep ~ arm, placebo = 'dose = 0.0',
                     data = locked_data, alternative = 'greater',
                     scale = 'risk difference')

  ## for summary of early termination
  trial$save(value = fit$z, name = 'z_value')
  trial$save(value = ifelse(fit$z > 1.64, 'add dose arms', 'stop trial'), 
             name = 'interim_decision')

  ## create three dose arms
  ep <- endpoint(name = 'ep', type = 'non-tte', readout = c(ep = 1),
                 generator = rbinom, size = 1, prob = .112)
  trt1 <- arm(name = 'dose = 0.5')
  trt1$add_endpoints(ep)

  ep <- endpoint(name = 'ep', type = 'non-tte', readout = c(ep = 1),
                 generator = rbinom, size = 1, prob = .208)
  trt2 <- arm(name = 'dose = 1.5')
  trt2$add_endpoints(ep)

  ep <- endpoint(name = 'ep', type = 'non-tte', readout = c(ep = 1),
                 generator = rbinom, size = 1, prob = .241)
  trt3 <- arm(name = 'dose = 2.5')
  trt3$add_endpoints(ep)

  ## add three new arms to trial
  trial$add_arms(sample_ratio = c(2, 2, 2), trt1, trt2, trt3)
  
}
```

<!-- Hide this action function and display it later for better typesetting. -->

```{r action_at_final, echo=FALSE}
action_at_final <- function(trial){

  locked_data <- trial$get_locked_data('final')

  trial$save(value = go_nogo(locked_data), name = 'decision')

}
```


## Define Milestones and Actions

Two milestones are triggered when 30 and 150 readouts are collected, 
respectively. Note that the 30 readouts are from the placebo and the highest 
dose arm, while the 150 readouts are from all five arms. 

```{r}
interim <- milestone(name = 'interim',
                     when = eventNumber(endpoint = 'ep', n = 30),
                     action = action_at_interim)

final <- milestone(name = 'final',
                   when = eventNumber(endpoint = 'ep', n = 150),
                   action = action_at_final)
```

In the action function `action_at_interim()`, we don't actually terminate the 
trial in simulation but always add dose arms. Instead, we save the z statistic 
for summary later, e.g., computing the proportion of early termination. 

```{r ref.label='action_at_interim', eval=FALSE}
```

At final analysis, we analyze data from all five arms using MCPMod. It calls 
a helper function `go_nogo()` which can be found in the Appendix below. 

```{r ref.label='action_at_final', eval=FALSE}
```

## Execute a Trial
After registering the two milestones with a listener, we simulate a single trial 
using `controller$run()`. We can see that more patients are randomized to the 
three newly added arms after interim, reflecting the new randomization ratio 
1:2:2:2:1. 

```{r message=FALSE}
listener <- listener()
listener$add_milestones(interim, final)

controller <- controller(trial, listener)
controller$run(n = 1, plot_event = TRUE)
```

In this trial, the z statistic at interim is relatively low (`r round(controller$get_output('z_value'), 2)`), thus is early 
terminated even if in simulation we keep going with three newly added dose arms.

```{r}
output <- controller$get_output()

output %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE,
                position = "left") %>%
  scroll_box(width = "100%")
```

We can call `controller$run(n = 10)` to simulate additional trials. 
in the function `action_at_final()`, the go/no-go decision (`decision` in output) 
is determined under the assumption that the dose has been expanded, 
regardless of the interim evidence—this simplification is made solely for the 
convenience of simulation. However, the overall probability of a `"go"` should 
incorporate the interim results—specifically, the trial must not be stopped at 
interim (i.e., `interim_decision == "add dose arms"`), 
and the function `go_nogo()` must return `"go"` at the final analysis. 

```{r message=FALSE}
## important: reset before calling run() again
controller$reset()
controller$run(n = 10, plot_event = FALSE, silent = TRUE)
res <- controller$get_output()

res %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE,
                position = "left") %>%
  scroll_box(width = "100%")
```

```{r message=FALSE}
## Number of overall Go
sum(res$interim_decision == 'add dose arms' & 'decision' == 'go')

## Number of early trial termination
sum(res$interim_decision == 'stop trial')
```

## Appendix: Code of Helper Function
For completeness, the full code of the helper functions `go_nogo()` is included 
below, which informs go/no-go decision at the end of the trial.

```{r ref.label='go_nogo', eval=FALSE}
```




