---
title: "An Example of Simulating a Trial with Adaptive Enrichment Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Example of Simulating a Trial with Adaptive Enrichment Design}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache.path = 'cache/enrichmentDesign/',
  comment = '#>',
  dpi = 300,
  out.width = '100%'
)
```

```{r setup, echo = FALSE, message = FALSE}
library(dplyr)
library(kableExtra)
library(survival)
library(ggplot2)
library(TrialSimulator)
```

In this vignette, we illustrate how to use `TrialSimulator` to simulate
a two-stage adaptive enrichment design. A binary baseline biomarker
splits the population into a biomarker-positive subgroup and its
complement. At an interim analysis, conditional power computed for the
full population and for the subgroup drives a five-zone decision: the
trial may stop for futility, continue unchanged, restrict further
enrollment to the subgroup (enrichment), or increase its number of
events and sample size (promising zone). The two stages are analyzed
separately and combined with the inverse normal method, and the
family-wise error rate over the full-population and subgroup hypotheses
is controlled by a closed test with Simes' test for the intersection
hypothesis.

The design borrows its building blocks from other vignettes: conditional
power and event number reassessment are described in
[Conditional power and event number reassessment](conditionalPower.html),
and the adaptation functions in
[Action function](actionFunctions.html).

## Simulation Settings

-   The trial consists of a placebo arm (`pbo`) and a treatment arm
    (`trt`). Patients are randomized with ratio `1:1`.
-   A binary baseline `biomarker` is observed at randomization. Its
    prevalence is 40%. Patients with `biomarker = 1` form the subgroup
    of interest.
-   The only efficacy endpoint is progression-free survival (`PFS`),
    which follows an exponential distribution whose median depends on
    the arm and on the biomarker. Medians are
    -   6 months in the placebo arm, in both biomarker strata;
    -   6.5 months in biomarker-negative patients of the treatment arm
        (hazard ratio 0.92);
    -   10 months in biomarker-positive patients of the treatment arm
        (hazard ratio 0.6).
-   6 patients are randomized per month for the first 6 months, and 10
    patients per month after then. The planned sample size is 150
    patients; it can be increased adaptively.
-   Dropout rate is 5% at month 12, modeled by an exponential
    distribution.
-   The trial has two stages. The first 80 randomized patients form the
    stage 1 cohort; patients randomized afterwards form the stage 2
    cohort. The two cohorts are analyzed separately and their p-values
    are combined with the inverse normal method with pre-specified
    weights $w_1 = \sqrt{65/110}$ and $w_2 = \sqrt{45/110}$, i.e., the
    planned split of the 110 events between the stages.
-   The final analysis is planned when 110 `PFS` events are observed in
    the whole trial, provided that the stage 1 cohort has reached 65
    events. Note that this is the condition as planned; it is replaced
    when the trial is adapted at the interim, as described below.
-   An interim decision is made when 45 `PFS` events are observed in
    the stage 1 cohort. Let $CP_{full}$ be the conditional power for the
    full population given the planned 110 events, and $CP_{sub}$ the
    conditional power for the subgroup given 90 events in the subgroup.
    Both are computed under the trend observed at the interim, at the
    one-sided level $\alpha = 0.025$ of the final test. The decision
    zones are
    -   *futility* if $CP_{full} < 0.01$ and $CP_{sub} < 0.01$: the
        trial stops at the interim, and neither hypothesis is rejected.
        The cutoff is chosen so that, under the alternative described
        above, the trial stops for futility in about 8% of the
        replicates;
    -   *unfavorable* if $CP_{full} < 0.2$ and $CP_{sub} < 0.6$ but
        not in the futility zone: the trial continues as planned;
    -   *enrichment* if $CP_{full} < 0.2$ and $CP_{sub} \ge 0.6$:
        enrollment is restricted to biomarker-positive patients, the
        accrual rate drops to 5 patients per month because of
        screening, up to 150 biomarker-positive patients are enrolled
        in total, and the number of events in the subgroup at the final
        analysis is re-estimated to reach a conditional power of 0.95,
        with a cap of 120 events and a floor of 50 events;
    -   *promising* if $0.2 \le CP_{full} < 0.9$: the sample size is
        increased to 240 patients and the total number of events at the
        final analysis is re-estimated to reach a conditional power of
        0.95, with a cap of 200 events;
    -   *favorable* if $CP_{full} \ge 0.9$: the trial continues as
        planned.
-   The stage 1 analysis is done when 65 `PFS` events are observed in
    the stage 1 cohort. Follow-up of the stage 1 cohort is stopped at
    that time, so that the stage 2 data are independent of the stage 1
    data. One-sided logrank p-values are computed for the full
    population ($p_{F,1}$) and for the subgroup ($p_{S,1}$), and Simes'
    p-value $p_{FS,1} = \min\{2\min(p_{F,1}, p_{S,1}), \max(p_{F,1}, p_{S,1})\}$
    for the intersection hypothesis.
-   At the final analysis, the same three p-values are computed from
    the stage 2 cohort. In the enrichment zone no biomarker-negative
    patient is enrolled in stage 2, so the full-population hypothesis
    cannot be tested there: its stage 2 p-value is set to 1 and the
    intersection p-value equals the subgroup p-value.
-   Each of the three p-values is combined across stages with the
    inverse normal method. The closed test rejects the full-population
    hypothesis if the combined intersection and full-population
    p-values are both at most 0.025 (one-sided), and rejects the
    subgroup hypothesis if the combined intersection and subgroup
    p-values are both at most 0.025.

## Define Two Arms

The biomarker and `PFS` are generated together by a custom generator,
so that the distribution of `PFS` can depend on the biomarker. A
generator returns a data frame with one column per endpoint. For a
time-to-event endpoint, an extra column with suffix `_event` holds the
event indicator, which is always 1 here because the generator produces
event times; censoring is applied by `TrialSimulator` at data locks and
adaptations. The biomarker is declared with `type = 'baseline'`, i.e., a
non-time-to-event endpoint observed at randomization, so it does not need a
readout time.

```{r feoaj, class.source="fold-show"}
rng <- function(n, median0, median1, prevalence){
  biomarker <- rbinom(n, size = 1, prob = prevalence)
  pfs <- rexp(n, rate = log(2) / ifelse(biomarker == 1, median1, median0))
  data.frame(biomarker = biomarker, pfs = pfs, pfs_event = 1)
}
```

The medians of the two arms are kept in two lists, because they are
needed again when the generators are updated at enrichment.

```{r oeiaj, class.source="fold-show"}
pbo_pars <- list(median0 = 6, median1 = 6)
trt_pars <- list(median0 = 6.5, median1 = 10)

pbo <- arm(name = 'pbo')
pbo$add_endpoints(
  endpoint(name = c('biomarker', 'pfs'), type = c('baseline', 'tte'),
           generator = rng, median0 = pbo_pars$median0,
           median1 = pbo_pars$median1, prevalence = .4))

trt <- arm(name = 'trt')
trt$add_endpoints(
  endpoint(name = c('biomarker', 'pfs'), type = c('baseline', 'tte'),
           generator = rng, median0 = trt_pars$median0,
           median1 = trt_pars$median1, prevalence = .4))
```

## Define a Trial

The trial starts with 150 planned patients and a piecewise constant
accrual rate. The end of the trial is controlled by milestones defined
below. A seed is set so that the single run shown later is reproducible;
in a formal simulation, leave `seed = NULL` and `TrialSimulator` picks
and records a seed for every replicate.

```{r ieoaj, class.source="fold-show"}
accrual_rate <- data.frame(end_time = c(6, Inf), piecewise_rate = c(6, 10))
trial <- trial(name = 'enrichment', n_patients = 150, seed = 13,
               enroller = StaggeredRecruiter, accrual_rate = accrual_rate,
               dropout = rexp, rate = -log(1 - .05) / 12, ## 5% by month 12
               silent = TRUE)
trial$add_arms(sample_ratio = c(1, 1), pbo, trt)
```

## Define Trial Milestones and Action Functions

The trial has three milestones, named `interim`, `stage 1 analysis` and
`final`. Their action functions are defined first, then the milestones
are registered.

### Interim decision

The action function of `interim` computes the two conditional powers
with `trial$conditionalPower()`. The subset condition `patient_id <= 80`
restricts the calculation to the stage 1 cohort, and `biomarker == 1`
further restricts it to the subgroup. The argument `D` is the number of
events assumed at the final analysis, `alpha` the one-sided level of the
final test, and `effect = 'trend'` uses the hazard ratio observed at the
interim.

In the futility zone, the trial stops. `TrialSimulator` has no dedicated
method for stopping a trial early: a trial ends when its last milestone
is triggered. Stopping is therefore expressed by
`trial$update_milestone()`, which moves the two remaining milestones to
the current calendar time, so that both are triggered right after the
interim and enrollment stops there. Their action functions, defined
below, skip the analyses when the decision is futility.

In the enrichment zone, four adaptations are made:

-   `trial$eventNumberReestimationFromConditionalPower()` finds the
    smallest number of events in the subgroup that reaches the target
    conditional power, subject to the cap `D_cap`;
-   `trial$resize()` raises the maximum sample size so that 150
    biomarker-positive patients are enrolled in total;
-   `trial$update_accrual_rate()` slows down accrual because screening
    for the biomarker takes time;
-   `trial$update_generator()` replaces the generators of both arms by
    the same generator with `prevalence = 1`, so that every patient
    enrolled from now on is biomarker-positive.

Finally, `trial$update_milestone()` replaces the triggering condition of
the final analysis by the re-estimated number of events in the
subgroup. In the promising zone, only the sample size and the total
number of events are updated. In both cases the condition keeps the
requirement of 65 events in the stage 1 cohort, so that the stage 1
analysis always happens before the final analysis.

```{r aoeif, class.source="fold-show"}
interim_action <- function(trial){

  locked_data <- trial$get_locked_data('interim')

  cp_full <- trial$conditionalPower(
    milestone = 'interim', Surv(pfs, pfs_event) ~ arm, placebo = 'pbo',
    alternative = 'less', alpha = .025, D = 110, effect = 'trend',
    patient_id <= 80)$cp

  cp_sub <- trial$conditionalPower(
    milestone = 'interim', Surv(pfs, pfs_event) ~ arm, placebo = 'pbo',
    alternative = 'less', alpha = .025, D = 90, effect = 'trend',
    patient_id <= 80 & biomarker == 1)$cp

  decision <- if(cp_full < .01 && cp_sub < .01){
    'futility'
  }else if(cp_full < .2 && cp_sub < .6){
    'unfavorable'
  }else if(cp_full < .2){
    'enrichment'
  }else if(cp_full < .9){
    'promising'
  }else{
    'favorable'
  }

  if(decision == 'futility'){

    ## stop the trial now: both remaining milestones are triggered at the
    ## current time, and their action functions skip the analyses
    now <- trial$get_current_time()
    trial$update_milestone(name = 'stage 1 analysis',
                           when = calendarTime(time = now))
    trial$update_milestone(name = 'final',
                           when = calendarTime(time = now))

  }else if(decision == 'enrichment'){

    en <- trial$eventNumberReestimationFromConditionalPower(
      milestone = 'interim', Surv(pfs, pfs_event) ~ arm, placebo = 'pbo',
      alternative = 'less', alpha = .025, target_cp = .95, effect = 'trend',
      patient_id <= 80 & biomarker == 1, D_cap = 120)

    max_events_in_subgroup <- ifelse(en$target_reached, en$D, 120)
    max_patients_in_subgroup <- 150
    enrolled_in_subgroup <- sum(locked_data$biomarker)

    trial$resize(n_patients = nrow(locked_data) +
                   max_patients_in_subgroup - enrolled_in_subgroup)
    trial$update_accrual_rate(
      accrual_rate = data.frame(end_time = Inf, piecewise_rate = 5))

    trial$update_generator(
      arm_name = 'pbo', endpoint_name = c('pfs', 'biomarker'),
      generator = rng, median0 = pbo_pars$median0,
      median1 = pbo_pars$median1, prevalence = 1.0)
    trial$update_generator(
      arm_name = 'trt', endpoint_name = c('pfs', 'biomarker'),
      generator = rng, median0 = trt_pars$median0,
      median1 = trt_pars$median1, prevalence = 1.0)

    trial$update_milestone(
      name = 'final',
      when = eventNumber(endpoint = 'pfs', n = max(max_events_in_subgroup, 50),
                         biomarker == 1) &
        eventNumber(endpoint = 'pfs', n = 65, patient_id <= 80))

  }else if(decision == 'promising'){

    en <- trial$eventNumberReestimationFromConditionalPower(
      milestone = 'interim', Surv(pfs, pfs_event) ~ arm, placebo = 'pbo',
      alternative = 'less', alpha = .025, target_cp = .95, effect = 'trend',
      patient_id <= 80, D_cap = 200)

    max_events <- ifelse(en$target_reached, en$D, 200)

    trial$resize(n_patients = 240)
    trial$update_milestone(
      name = 'final',
      when = eventNumber(endpoint = 'pfs', n = max_events) &
        eventNumber(endpoint = 'pfs', n = 65, patient_id <= 80))

  }
  ## 'unfavorable' and 'favorable': no adaptation

  trial$save(value = decision, name = 'decision')
  trial$save(value = cp_full, name = 'cp_full')
  trial$save(value = cp_sub, name = 'cp_subgroup')
}
```

### Stage 1 analysis

At `stage 1 analysis`, `trial$stop_followup(patient_id <= 80)` stops
the follow-up of the stage 1 cohort at the milestone time. Data of these
patients are therefore identical in every later data lock, which is what
makes the stage 2 p-values independent of the stage 1 p-values. The
three stage 1 p-values are saved with `trial$save()` for use at the
final analysis. Note that subset conditions such as `patient_id <= 80`
are accepted by `fitLogrank()` too. When the trial has stopped for
futility, the p-values are saved as `NA` so that every replicate has the
same output columns.

```{r iefoa, class.source="fold-show"}
stage1_action <- function(trial){

  if(trial$get_output('decision') == 'futility'){
    trial$save(value = NA, name = 'pf1')
    trial$save(value = NA, name = 'ps1')
    trial$save(value = NA, name = 'simes1')
    return(invisible(NULL))
  }

  trial$stop_followup(patient_id <= 80)

  locked_data <- trial$get_locked_data('stage 1 analysis')

  pf1 <- fitLogrank(Surv(pfs, pfs_event) ~ arm, placebo = 'pbo',
                    data = locked_data, alternative = 'less',
                    patient_id <= 80)$p
  ps1 <- fitLogrank(Surv(pfs, pfs_event) ~ arm, placebo = 'pbo',
                    data = locked_data, alternative = 'less',
                    patient_id <= 80 & biomarker == 1)$p

  trial$save(value = pf1, name = 'pf1')
  trial$save(value = ps1, name = 'ps1')
  trial$save(value = min(2 * min(pf1, ps1), max(pf1, ps1)), name = 'simes1')
}
```

### Final analysis

At the final analysis, values saved earlier in the same replicate are
retrieved with `trial$get_output()`. The stage 2 p-values use the stage
2 cohort only (`patient_id > 80`). The combination and the closed test
are then a few lines of arithmetic.

```{r aoeigj, class.source="fold-show"}
final_action <- function(trial){

  locked_data <- trial$get_locked_data('final')
  decision <- trial$get_output('decision')

  trial$save(value = sum(locked_data$biomarker), name = 'n_positive')

  if(decision == 'futility'){
    trial$save(value = NA, name = 'pf2')
    trial$save(value = NA, name = 'ps2')
    trial$save(value = NA, name = 'simes2')
    trial$save(value = FALSE, name = 'reject_full')
    trial$save(value = FALSE, name = 'reject_subgroup')
    return(invisible(NULL))
  }

  pf1 <- trial$get_output('pf1')
  ps1 <- trial$get_output('ps1')
  simes1 <- trial$get_output('simes1')

  if(decision == 'enrichment'){
    ps2 <- fitLogrank(Surv(pfs, pfs_event) ~ arm, placebo = 'pbo',
                      data = locked_data, alternative = 'less',
                      patient_id > 80 & biomarker == 1)$p
    pf2 <- 1
    simes2 <- ps2
  }else{
    pf2 <- fitLogrank(Surv(pfs, pfs_event) ~ arm, placebo = 'pbo',
                      data = locked_data, alternative = 'less',
                      patient_id > 80)$p
    ps2 <- fitLogrank(Surv(pfs, pfs_event) ~ arm, placebo = 'pbo',
                      data = locked_data, alternative = 'less',
                      patient_id > 80 & biomarker == 1)$p
    simes2 <- min(2 * min(pf2, ps2), max(pf2, ps2))
  }

  trial$save(value = pf2, name = 'pf2')
  trial$save(value = ps2, name = 'ps2')
  trial$save(value = simes2, name = 'simes2')

  w1 <- sqrt(65 / 110)
  w2 <- sqrt(45 / 110)
  combine <- function(p1, p2){
    1 - pnorm(w1 * qnorm(1 - p1) + w2 * qnorm(1 - p2))
  }
  pf <- combine(pf1, pf2)
  ps <- combine(ps1, ps2)
  simes <- combine(simes1, simes2)

  trial$save(value = (simes <= .025 && pf <= .025), name = 'reject_full')
  trial$save(value = (simes <= .025 && ps <= .025), name = 'reject_subgroup')
}
```

### Register milestones

The three milestones are all event-driven. The conditions of `interim`
and `stage 1 analysis` count events in the stage 1 cohort only, through
the subset condition `patient_id <= 80`. The condition of `final` is the
one planned at design; it is replaced in the action function of
`interim` when the trial is adapted.

```{r eaofj, class.source="fold-show"}
interim <- milestone(name = 'interim', action = interim_action,
                     when = eventNumber(endpoint = 'pfs', n = 45, patient_id <= 80))
stage1 <- milestone(name = 'stage 1 analysis', action = stage1_action,
                    when = eventNumber(endpoint = 'pfs', n = 65, patient_id <= 80))
final <- milestone(name = 'final', action = final_action,
                   when = eventNumber(endpoint = 'pfs', n = 110) &
                     eventNumber(endpoint = 'pfs', n = 65, patient_id <= 80))

listener <- listener(silent = TRUE)
listener$add_milestones(interim, stage1, final)
```

## Execute a Trial

With the seed set above, the interim falls in the enrichment zone. The
event plot shows cumulative counts by arm over calendar time, with the
three milestones as dashed lines. The panel of `biomarker`, a baseline
endpoint, is the enrollment curve: it bends at the interim when accrual
slows down to biomarker-positive patients only, and reaches the
increased sample size well before the final analysis, which waits for
the re-estimated number of events in the subgroup.

```{r oeaif, dpi = 1200}
controller <- controller(trial, listener)
controller$run(plot_event = TRUE, silent = TRUE)
```

The output of the replicate collects the milestone times, the number of
patients and events at each milestone, and every value saved in the
action functions. In this replicate, the subgroup hypothesis is rejected
at the final analysis, while the full-population hypothesis is not, as
it cannot be under enrichment.

```{r iaefj, class.source="fold-show"}
controller$get_output() %>%
  kable(escape = FALSE) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  scroll_box(width = "100%")
```

## Execute Trial Simulation

Operating characteristics are obtained by running many replicates. The
code below was run with 10000 replicates under the alternative described
in the settings, and another 10000 replicates under the global null where
the median of `PFS` in the treatment arm equals the placebo median of 6
months in both biomarker strata. The results are stored in the package.

```{r eoaif, eval=FALSE}
## reset a controller if $run has been executed before
controller$reset()
controller$run(n = 10000, plot_event = FALSE, silent = TRUE, tidy = TRUE)
output <- controller$get_output()
```

```{r oaeif, echo=FALSE}
output <- TrialSimulator:::getEnrichmentDesignOutput()
zones <- c('futility', 'unfavorable', 'enrichment', 'promising', 'favorable')
output$decision <- factor(output$decision, levels = zones)
```

```{r iaeof, class.source = "fold-hide"}
output %>%
  filter(scenario == 'alternative') %>%
  select(-scenario) %>%
  head(5) %>%
  kable(escape = FALSE) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  scroll_box(width = "100%")
```

Under the alternative, the five zones are reached with the following
frequencies, and the adaptations change the duration and the size of the
trial accordingly. In the futility zone, the trial ends at the interim,
so the three milestone times coincide.

```{r aoeij, class.source = "fold-hide"}
output %>%
  filter(scenario == 'alternative') %>%
  group_by(decision) %>%
  summarise(
    zone = n() / 100,
    time_interim = mean(`milestone_time_<interim>`),
    time_stage1 = mean(`milestone_time_<stage 1 analysis>`),
    time_final = mean(`milestone_time_<final>`),
    n_final = mean(`n_events_<final>_<patient_id>`),
    n_positive_final = mean(n_positive),
    events_final = mean(`n_events_<final>_<pfs>`)
  ) %>%
  kable(col.names = c('Zone', 'Zone (%)', 'Interim', 'Stage 1', 'Final',
                      'Patients', 'Positive Patients', 'PFS Events'),
        digits = 1, align = 'r',
        caption = 'Decision Zones, Milestone Times, and Trial Size at the Final Analysis') %>%
  add_header_above(c(' ' = 2, 'Time (Months)' = 3, 'At Final Analysis' = 3)) %>%
  kable_styling(full_width = TRUE)
```

Power is summarized by zone and overall. The last column is the
probability of rejecting at least one of the two hypotheses.

```{r eiaof, class.source = "fold-hide"}
power_table <- function(dat){
  dat %>%
    summarise(
      zone = n() / 100,
      reject_full_only = 100 * mean(reject_full & !reject_subgroup),
      reject_subgroup_only = 100 * mean(!reject_full & reject_subgroup),
      reject_both = 100 * mean(reject_full & reject_subgroup),
      reject_any = 100 * mean(reject_full | reject_subgroup)
    )
}

alt <- output %>% filter(scenario == 'alternative')
bind_rows(
  alt %>% group_by(decision) %>% power_table(),
  alt %>% power_table() %>% mutate(decision = 'overall')
) %>%
  kable(col.names = c('Zone', 'Zone (%)', 'Full Only', 'Subgroup Only',
                      'Both', 'Any'),
        digits = 1, align = 'r',
        caption = 'Power (%) under the Alternative') %>%
  add_header_above(c(' ' = 2, 'Rejected Hypotheses (%)' = 4)) %>%
  kable_styling(full_width = TRUE)
```

Under the global null, the probability of rejecting at least one
hypothesis is the family-wise error rate. The closed test with the
inverse normal combination controls it at the one-sided level 0.025 in
the strong sense, whatever adaptation is chosen at the interim. Note
that the rates within a zone are conditional on the interim decision:
they are not controlled individually, and they are large in the
favorable zone precisely because that zone is reached when the interim
data look good by chance. Only the overall rate matters. With 10000
replicates, the Monte Carlo standard error of the overall rate is about
0.16 percentage points.

```{r oeiaf, class.source = "fold-hide"}
null <- output %>% filter(scenario == 'null')
bind_rows(
  null %>% group_by(decision) %>% power_table(),
  null %>% power_table() %>% mutate(decision = 'overall')
) %>%
  kable(col.names = c('Zone', 'Zone (%)', 'Full Only', 'Subgroup Only',
                      'Both', 'Any'),
        digits = 1, align = 'r',
        caption = 'Type I Error (%) under the Global Null') %>%
  add_header_above(c(' ' = 2, 'Rejected Hypotheses (%)' = 4)) %>%
  kable_styling(full_width = TRUE)
```
