---
title: "An Example of Fixed Design with Two Correlated Endpoints"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Example of Fixed Design with Two Correlated Endpoints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache.path = 'cache/fixedDesign/',
  comment = '#>',
  dpi = 300,
  out.width = '100%'
)
```

```{r setup, echo = FALSE, message = FALSE}
library(dplyr)
library(kableExtra)
library(TrialSimulator)
set.seed(12345)
```

In this vignette, we illustrate how to use `TrialSimulator` to simulate a trial 
of two correlated endpoints. Under a fixed design, only one milestone is specified 
for final analysis. 

## Simulation Settings

-   Trial consists of two active arms of high and low dose, 
    and a standard-of-care arm.
-   Patients are randomized into the trial with ratio `1:1:1`.
-   30 patients are randomized per month for the first 10 months, and 50
    patients per month after then until 1000 patients.
-   Dropout rate is 10% at month 18, which is modeled by an exponential 
    distribution, with rate parameter `-log(1 - 0.1)/18`. 

-   Modeled by a three-state ill-death model, two endpoints `PFS` and `OS` 
    are simulated.
    -   Primary endpoint `OS` has a medians of 15, 18.5, 20 months in standard-of-care, 
        low dose and high dose arms, respectively. 
    -   Key secondary endpoint `PFS` has a median of 7, 9, 10 months in the 
        three arms. 
    -   Pearson's correlation between `OS` and `PFS` are 0.68, 0.65, and 0.60 in 
        the three arms. 

-   To ensures sufficient powers for both endpoints, final analysis is set when 
    we have at least a total of 800 `PFS` events in the standard-of-care and 
    high dose arm, meanwhile a total of 550 `OS` events are observed in three 
    arms
    -   `OS` is tested using one-sided logrank test. 
    -   `PFS` is tested using one-sided p-value from the Cox proportional hazard 
        model as the PH assumption is assumed. 
    -   The Bonferroni correction is adopted to split overall $\alpha = 5\%$, 
        i.e., claiming significant effect for an endpoint when p-value is lower 
        than $0.05/4$. 
    
## Transition Hazards of `PFS` and `OS`

We adopt the ill-death model to simulate the two endpoints. This ensures `PFS` 
$\leq$ `OS` with probability one, and makes no assumption on latent variables or 
copula parameters. `TrialSimulator` offers a function `solveThreeStateModel()` 
to convert endpoints' medians and correlation to the transition hazards, which 
are required by the built-in generator `CorrelatedPfsAndOs3()`. Refer to the 
vignette [Simulate Correlated Progression-Free Survival and Overall Survival as Endpoints in Clinical Trials](simulatePfsAndOs.html) for more details. 

```{r iioeal}
pars_soc <- solveThreeStateModel(median_pfs = 7, median_os = 15, corr = .68, 
                                 h12 = seq(.07, .10, length.out = 50))
pars_soc
```

```{r eaioie}
pars_low <- solveThreeStateModel(median_pfs = 9, median_os = 18.5, corr = .65, 
                                 h12 = seq(.04, .07, length.out = 50))
pars_low
```

```{r aealga}
pars_high <- solveThreeStateModel(median_pfs = 10, median_os = 20, corr = .60, 
                                  h12 = seq(.02, .06, length.out = 50))
pars_high
```

```{r eaald, echo=FALSE, eval=TRUE}
rbind(pars_soc, pars_low, pars_high) %>% 
  structure(class = 'data.frame') %>% 
  mutate(arm = c('soc', 'low', 'high')) %>% 
  select(arm, h01, h02, h12) %>% 
  kable(format = 'html', digits = 3, 
        caption = 'Transition hazards in three treatment arms')
```

## Define Treatment Arms

We use generator `CorrelatedPfsAndOs3()` with `endpoint()` and `arm()` to define 
the three treatment arms. 

```{r ieaof}
#' define SoC
pfs_os_in_soc <- endpoint(name = c('pfs', 'os'), 
                          type = c('tte', 'tte'), 
                          generator = CorrelatedPfsAndOs3, 
                          h01 = 0.075, h02 = 0.024, h12 = 0.090)

soc <- arm(name = 'soc')
soc$add_endpoints(pfs_os_in_soc)

#' define low dose arm
pfs_os_in_low <- endpoint(name = c('pfs', 'os'), 
                          type = c('tte', 'tte'), 
                          generator = CorrelatedPfsAndOs3, 
                          h01 = 0.051, h02 = 0.026, h12 = 0.062)

low <- arm(name = 'low')
low$add_endpoints(pfs_os_in_low)

#' define high dose arm
pfs_os_in_high <- endpoint(name = c('pfs', 'os'), 
                           type = c('tte', 'tte'), 
                           generator = CorrelatedPfsAndOs3, 
                          h01 = 0.040, h02 = 0.030, h12 = 0.047)

high <- arm(name = 'high')
high$add_endpoints(pfs_os_in_high)
```

We can request for a summary report of, e.g., the high dose arm, by printing 
the arm object in R console. The medians of `PFS` and `OS` matches to the settings 
very well. 

```{r ioeafaea, results='asis'}
high
```

## Define a Trial

With three arms, we can define a trial using the function `trial()`. Recruitment curve are specified through `enroller` with a built-in function `StaggeredRecruiter` of piecewise constant rate. We set `duration` to be an arbitrary large number (500) but controlling the end of trial through a pre-defined milestone later. Note that if `seed = NULL`, `TrialSimulator` will pick a seed for the purpose of reproducibility.

```{r ioeeaf}
accrual_rate <- data.frame(end_time = c(10, Inf),
                           piecewise_rate = c(30, 50))
trial <- trial(
  name = 'Trial-3415', n_patients = 1000,
  seed = 1727811904, duration = 500,
  enroller = StaggeredRecruiter, accrual_rate = accrual_rate,
  dropout = rexp, rate = -log(1 - 0.1)/18, ## 10% by month 18
  silent = TRUE
)

trial$add_arms(sample_ratio = c(1, 1, 1), soc, low, high) ## 1:1:1
trial
```


## Define Milestone and Action for Final Analysis

To ensure sufficient powers of testing `PFS` and `OS`, final analysis is performed 
when we have at least 700 events for `OS` and 800 events for `PFS`. In the 
action function, we compute one-sided p-value of `PFS` using the proportional 
hazard Cox model, and one-sided p-value of `OS` using the logrank test. This is 
consistent with the assumption of ill-death model implemented in data generator 
`CorrelatedPfsAndOs3()`. Note that five columns are available in locked data: 
`arm`, `pfs`, 'os', 'pfs_event', and 'os_event', which are used to construct 
model formula. Estimates of hazard ratio are also computed. Built-in functions 
`fitCoxph` and `fitLogrank` return data frames. Refer to their help documants 
for more details. 

```{r wewfal}
action <- function(trial){
  
  locked_data <- trial$get_locked_data('final')
  
  pfs <- fitCoxph((Surv(pfs, pfs_event) ~ arm), placebo = 'soc', 
                  data = locked_data, alternative = 'less', 
                  scale = 'hazard ratio')
  
  os <- fitLogrank((Surv(os, os_event) ~ arm), placebo = 'soc', 
                   data = locked_data, alternative = 'less')
  
  ## Bonferroni test is applied to four hypotheses: 
  ## PFS_low, PFS_high, OS_low, and OS_high
  pfs$decision <- ifelse(pfs$p < .05/4, 'reject', 'accept')
  os$decision <- ifelse(os$p < .05/4, 'reject', 'accept')
  
  trial$save(
    value = pfs %>% filter(arm == 'low') %>% select(estimate, decision, info), 
    name = 'pfs_low')
  
  trial$save(
    value = pfs %>% filter(arm == 'high') %>% select(estimate, decision, info), 
    name = 'pfs_high')
  
  trial$save(
    value = os %>% filter(arm == 'low') %>% select(decision, info), 
    name = 'os_low')
  
  trial$save(
    value = os %>% filter(arm == 'high') %>% select(decision, info), 
    name = 'os_high')
  
}
```

Now we can define and register the milestone to a listener, which monitors 
the trial for us through a controller

```{r tuyo}
final <- milestone(name = 'final', action = action, 
                   when = eventNumber(endpoint = 'pfs', n = 450, 
                                      arms = c('soc', 'high')) & 
                     eventNumber(endpoint = 'os', n = 550)
                   )

listener <- listener()
listener$add_milestones(final)

controller <- controller(trial, listener)
```

We can run a massive number of replicates in simulation to study
operating characteristics of a trial design by specifying `n` in
`Controller$run()`. We can set `plot_event = FALSE` to turn off plotting
to save running time. The simulation results can be accessed by calling the 
member function `get_output()` of the controller. 

```{r hgajffal, eval=FALSE, cache=TRUE, message=FALSE, warning=FALSE, results='hide'}
controller$run(n = 1000, plot_event = FALSE, silent = TRUE)
output <- controller$get_output()
```


```{r eaioalafj, echo=FALSE}
output <- TrialSimulator:::getFixedDesignOutput()
```

```{r eeioafla}
output %>% 
  head(5) %>% 
  kable(escape = FALSE) %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE,
                position = "left") %>%
  scroll_box(width = "100%")
```

For example, we can compute the powers and summarize the estimates of 
hazard ratio for `PFS`. 

```{r hghgale}
output %>% 
  summarise(
    across(matches('_<decision>$'), ~ mean(. == 'reject') * 100, .names = 'Power_{.col}'), 
    across(matches('_<estimate>$'), ~ mean(.x), .names = 'HR_{.col}')
  ) %>%
  rename_with(~ sub('_<decision>$', '', .), starts_with('Power_')) %>%
  rename_with(~ sub('_<estimate>$', '', .), starts_with('HR_')) %>% 
  kable(col.name = NULL, digits = 3, align = 'r') %>% 
  add_header_above(c('Low', 'High', 'Low', 'High', 'Low', 'High'), align = 'r') %>% 
  add_header_above(c('PFS' = 2, 'OS' = 2, 'PFS' = 2)) %>% 
  add_header_above(c('Power (%)' = 4, 'Hazard Ratio' = 2)) %>% 
  kable_styling(full_width = TRUE)
```

Note that `OS` does not satisfy the proportional hazard assumption, and a 
composite condition on event numbers is used to trigger the final analysis. 
Thus, the powers in the table above would not match to the output from power 
calculation packages, which is as expected. 

