---
title: "Simulate Correlated Progression-Free Survival and Overall Survival Using a Gumbel Copula"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate Correlated Progression-Free Survival and Overall Survival Using a Gumbel Copula}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache.path = 'cache/simulatePfsAndOsGumbel/',
  comment = '#>',
  dpi = 300,
  out.width = '100%'
)
```

```{r setup, echo = FALSE, message = FALSE}
library(dplyr)
library(knitr)
library(ggplot2)
library(TrialSimulator)

if (!requireNamespace('copula', quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

Progression-free survival (PFS) and overall survival (OS) are common time-to-event endpoints in oncology trials. Because progression or death defines a PFS event, simulated trial data should always satisfy $\mbox{PFS} \leq \mbox{OS}$. At the same time, the two endpoints are often highly correlated, so independent marginal simulation is usually not appropriate.

This vignette describes the copula-based generator implemented in `TrialSimulator::CorrelatedPfsAndOs2()`. The function is intended for simulation settings where the user wants to specify three interpretable quantities:

- the marginal median PFS,
- the marginal median OS,
- Kendall's tau between the observed, uncensored PFS and OS times.

The method uses a Gumbel survival copula for a latent time to progression (TTP) and OS, and then defines

$$
\mbox{PFS} = \min(\mbox{TTP}, \mbox{OS}).
$$

This construction guarantees $\mbox{PFS} \leq \mbox{OS}$ for every simulated patient. It also keeps both marginal PFS and OS exponential, so the inputs can be specified directly through their medians.

This is a useful distinction from the illness-death model in `TrialSimulator::CorrelatedPfsAndOs3()`. The illness-death model gives a clinically interpretable transition structure, but OS generated from that model can have a time-varying hazard ratio between treatment arms. Therefore, if OS will be analyzed using a proportional hazards Cox model and the simulation should be consistent with that analysis model, the copula-based method described here is often more appropriate.

## Model

Let $X$ denote latent TTP and let $Y$ denote OS. The model assumes exponential marginal survival functions

$$
\Pr(X > x) = \exp(-\lambda_X x), \qquad
\Pr(Y > y) = \exp(-\lambda_Y y),
$$

and a Gumbel--Hougaard survival copula

$$
\Pr(X > x, Y > y)
=
\exp\left[
-\left\{(\lambda_X x)^\theta + (\lambda_Y y)^\theta\right\}^{1/\theta}
\right],
\qquad \theta \geq 1.
$$

The observed PFS time is

$$
P = \min(X, Y).
$$

Under this model, OS is exponential with rate $\lambda_Y$. PFS is also exponential, with survival function

$$
\Pr(P > t)
=
\Pr(X > t, Y > t)
=
\exp\left[
-\left\{(\lambda_X^\theta + \lambda_Y^\theta)t^\theta\right\}^{1/\theta}
\right]
=
\exp(-\lambda_P t),
$$

where

$$
\lambda_P = \left(\lambda_X^\theta + \lambda_Y^\theta\right)^{1/\theta}.
$$

Therefore, if the target medians are $m_P$ for PFS and $m_Y$ for OS,

$$
\lambda_P = \frac{\log(2)}{m_P},
\qquad
\lambda_Y = \frac{\log(2)}{m_Y}.
$$

For a candidate value of $\theta$, the latent TTP rate is then

$$
\lambda_X =
\left(\lambda_P^\theta - \lambda_Y^\theta\right)^{1/\theta}.
$$

This requires $m_P < m_Y$, as expected for PFS and OS.

## Choosing the Copula Parameter

For the Gumbel copula, Kendall's tau between latent TTP and OS is

$$
\tau(X,Y) = 1 - \frac{1}{\theta}.
$$

However, `CorrelatedPfsAndOs2()` asks for Kendall's tau between the observed, uncensored endpoints $P=\min(X,Y)$ and $Y$, not between latent TTP and OS. These are not the same quantity. The formula below is derived in the appendix. Under the model above,

$$
\tau(P,Y)
=
1 - \frac{1}{\theta}
\left[
1 - \left(\frac{m_P}{m_Y}\right)^\theta
\right].
$$

Equivalently, writing $\tau_X = \tau(X,Y)$ and $\theta = 1/(1-\tau_X)$,

$$
\tau(P,Y)
=
1 -
(1-\tau_X)
\left[
1 -
\left(\frac{m_P}{m_Y}\right)^{1/(1-\tau_X)}
\right].
$$

`CorrelatedPfsAndOs2()` solves this equation numerically for the latent Kendall's tau $\tau_X$, converts it to $\theta$, and then simulates from the Gumbel copula.

For any finite $\theta$, $\tau(P,Y)$ is larger than $\tau(X,Y)$:

$$
\tau(P,Y)-\tau(X,Y)
=
\frac{1}{\theta}
\left(\frac{m_P}{m_Y}\right)^\theta
> 0.
$$

Intuitively, even if latent TTP and OS have a weaker association, observed PFS contains deaths by construction, which increases the association between observed PFS and OS.

The requested value of $\tau(P,Y)$ has a lower bound. When $\theta=1$, TTP and OS are independent, but PFS still contains OS deaths through the definition $P=\min(X,Y)$. In this limiting case,

$$
\tau(P,Y) = \frac{m_P}{m_Y}.
$$

Thus, for fixed medians $m_P$ and $m_Y$, `CorrelatedPfsAndOs2()` can only target

$$
\frac{m_P}{m_Y} \leq \tau(P,Y) < 1.
$$

If the requested Kendall's tau is too small for the two medians, the function stops with an error.

## Example

Suppose we want to simulate PFS with median 5 months and OS with median 11 months, targeting Kendall's tau of 0.6 between observed, uncensored PFS and OS times.

```{r example-endpoint, results = 'asis'}
pfs_and_os <- endpoint(name = c('pfs', 'os'),
                       type = c('tte', 'tte'),
                       generator = CorrelatedPfsAndOs2,
                       median_pfs = 5,
                       median_os = 11,
                       kendall = 0.6,
                       pfs_name = 'pfs',
                       os_name = 'os')

pfs_and_os
```

For verification only, we can call the generator directly. This is not the recommended way to use `TrialSimulator` for simulation studies; in practice, the generator should usually be supplied to `endpoint()` and used inside the trial simulation workflow. Direct calls are helpful here because they let us check the marginal medians, the observed Kendall's tau, and the ordering $\mbox{PFS} \leq \mbox{OS}$ before adding enrollment, censoring, treatment arms, and analyses.

```{r direct-generator}
set.seed(123)
dat <- CorrelatedPfsAndOs2(n = 100000,
                           median_pfs = 5,
                           median_os = 11,
                           kendall = 0.6)

head(dat, 2)
```

The simulation should approximately recover the requested medians and Kendall's tau between observed, uncensored PFS and OS times.

```{r validation}
with(dat, median(pfs))
with(dat, median(os))
with(dat, cor(pfs, os, method = 'kendall'))
with(dat, all(pfs <= os))
```

Because this generator returns event indicators equal to 1 for both endpoints, censoring and staggered enrollment should be handled by the broader trial simulation workflow rather than by this endpoint generator.

## Practical Guidance

This generator is most useful when the simulation objective is to specify marginal PFS and OS medians and a rank-based association between the two observed endpoints. Kendall's tau is often more stable and interpretable than Pearson correlation for skewed time-to-event variables, and it is directly connected to the Gumbel copula parameter.

There are still modeling assumptions to keep in mind:

- PFS and OS are both marginally exponential.
- Dependence is induced through a Gumbel survival copula on latent TTP and OS.
- The Gumbel copula only allows non-negative dependence.
- The requested Kendall's tau is for observed PFS and OS event times before censoring is applied.
- The attainable Kendall's tau between observed PFS and OS is constrained by the PFS/OS median ratio.

For simulations that require transition-specific hazards or a mechanistic disease process, the illness-death model implemented in `TrialSimulator::CorrelatedPfsAndOs3()` may be more appropriate. For simulations where the main requirement is a direct medians-plus-Kendall's-tau specification, especially when OS will be analyzed under a proportional hazards Cox model, `CorrelatedPfsAndOs2()` provides a compact alternative.

## Appendix: Kendall's Tau for Observed PFS and OS

This appendix gives the derivation behind the formula used in `CorrelatedPfsAndOs2()`.

Let $X$ and $Y$ be nonnegative random variables with exponential margins

$$
X \sim \operatorname{Exp}(\lambda_X),
\qquad
Y \sim \operatorname{Exp}(\lambda_Y),
$$

and joint survival function

$$
S_{X,Y}(x,y)
=
\Pr(X>x, Y>y)
=
\exp\left[
-\left\{(\lambda_X x)^\theta + (\lambda_Y y)^\theta\right\}^{1/\theta}
\right],
\qquad \theta \ge 1.
$$

Define $P=\min(X,Y)$. We want Kendall's tau between $P$ and $Y$.

Standardize the margins by setting

$$
A=\lambda_X X, \qquad B=\lambda_Y Y.
$$

Then the joint survival function of $(A,B)$ is

$$
\Pr(A>a, B>b)
=
\exp\left\{-(a^\theta+b^\theta)^{1/\theta}\right\}.
$$

Use the polar-type transformation

$$
R=(A^\theta+B^\theta)^{1/\theta},
\qquad
W=\frac{A^\theta}{A^\theta+B^\theta}.
$$

A Jacobian calculation gives the joint density

$$
f_{R,W}(r,w)=\frac{1}{\theta} e^{-r}(r+\theta-1),
\qquad r>0,\; 0<w<1.
$$

Thus, $R$ and $W$ are independent and $W \sim \operatorname{Unif}(0,1)$. The inverse transformation is

$$
A = R W^{1/\theta},
\qquad
B = R(1-W)^{1/\theta}.
$$

Let

$$
a=\lambda_X^\theta,
\qquad
b=\lambda_Y^\theta,
\qquad
c=\frac{a}{a+b},
\qquad
d=\frac{b}{a+b}.
$$

The event $X \le Y$ is equivalent to $W \le c$:

$$
X \le Y
\iff
\frac{A}{\lambda_X} \le \frac{B}{\lambda_Y}
\iff
W \le c,
$$

so

$$
P=
\begin{cases}
X, & W \le c,\\
Y, & W > c.
\end{cases}
$$

Using the survival-copula representation of Kendall's tau,

$$
\tau(P,Y)=4E\{S_{P,Y}(P,Y)\}-1.
$$

When $W \le c$, $P=X$ and

$$
S_{P,Y}(P,Y)=S_{X,Y}(X,Y)=e^{-R}.
$$

When $W>c$, $P=Y$. Since

$$
S_{P,Y}(y,y)=\Pr(P>y,Y>y)=\Pr(X>y,Y>y),
$$

we have, using $Y=B/\lambda_Y=R(1-W)^{1/\theta}/\lambda_Y$,

$$
S_{P,Y}(Y,Y)
=
\exp\left[
-R\left(\frac{1-W}{d}\right)^{1/\theta}
\right].
$$

Therefore,

$$
E\{S_{P,Y}(P,Y)\}
=
E\{e^{-R}\mathbf 1(W\le c)\}
+
E\left[
e^{-R((1-W)/d)^{1/\theta}}
\mathbf 1(W>c)
\right].
$$

By independence of $R$ and $W$,

$$
E\{S_{P,Y}(P,Y)\}
=
cL_R(1)
+
\int_c^1
L_R\left\{\left(\frac{1-w}{d}\right)^{1/\theta}\right\}
dw,
$$

where $L_R(s)=E(e^{-sR})$. From the density of $R$,

$$
L_R(s)
=
\frac{1}{\theta}
\int_0^\infty e^{-(1+s)r}(r+\theta-1)\,dr
=
\frac{\theta+(\theta-1)s}{\theta(1+s)^2}.
$$

In particular,

$$
L_R(1)=\frac{2\theta-1}{4\theta}.
$$

For the integral term, use the substitution

$$
t=\left(\frac{1-w}{d}\right)^{1/\theta},
\qquad
dw=-d\theta t^{\theta-1}dt.
$$

Then

$$
\int_c^1
L_R\left\{\left(\frac{1-w}{d}\right)^{1/\theta}\right\}
dw
=
d\theta\int_0^1 t^{\theta-1}L_R(t)\,dt.
$$

Substituting $L_R(t)$ gives

$$
d\int_0^1
\frac{\theta t^{\theta-1}+(\theta-1)t^\theta}{(1+t)^2}
dt.
$$

Because

$$
\frac{d}{dt}\left(\frac{t^\theta}{1+t}\right)
=
\frac{\theta t^{\theta-1}+(\theta-1)t^\theta}{(1+t)^2},
$$

the integral is

$$
d\left[\frac{t^\theta}{1+t}\right]_0^1
=
\frac{d}{2}.
$$

Thus,

$$
E\{S_{P,Y}(P,Y)\}
=
c\frac{2\theta-1}{4\theta}+\frac{d}{2}
=
\frac{2\theta-1+d}{4\theta},
$$

using $c+d=1$. Finally,

$$
\tau(P,Y)
=
4E\{S_{P,Y}(P,Y)\}-1
=
\frac{\theta-1+d}{\theta}
=
1-\frac{1}{\theta}
\left(
\frac{\lambda_X^\theta}{\lambda_X^\theta+\lambda_Y^\theta}
\right).
$$

Since $P$ has rate

$$
\lambda_P =
\left(\lambda_X^\theta+\lambda_Y^\theta\right)^{1/\theta},
$$

this can also be written as

$$
\tau(P,Y)
=
1-\frac{1}{\theta}
\left(
\frac{\lambda_X^\theta}{\lambda_P^\theta}
\right)
=
1-\frac{1}{\theta}
\left[
1-\left(\frac{\lambda_Y}{\lambda_P}\right)^\theta
\right].
$$

Because $\lambda_Y/\lambda_P = m_P/m_Y$, the formula used by `CorrelatedPfsAndOs2()` is

$$
\tau(P,Y)
=
1-\frac{1}{\theta}
\left[
1-\left(\frac{m_P}{m_Y}\right)^\theta
\right].
$$
